---
title: "Discrete random variables"
author: "[MACS 33000](https://jmclip.github.io/MACSS_math_camp/) <br /> University of Chicago"
output:
  xaringan::moon_reader:
    css: ["default", "xaringan-themer.css", "slides-menu.css"]
    math: katex
    includes:
      after_body: [slides-menu.js]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false 
---

```{r child = here::here("_common-slides.Rmd"), cache = FALSE}

```

```{r packages, include=FALSE}
library(tidyverse)
library(patchwork)
library(here)
library(gganimate)
```

$$\newcommand{\E}{\mathrm{E}} \newcommand{\Var}{\mathrm{Var}} \newcommand{\Cov}{\mathrm{Cov}}$$
# Misc

* CALCULATORS (10 total) -- practice questions submitted and calculators requested here: https://forms.gle/veuU3LehJdH8oxXT9 
* Survey: https://forms.gle/gE1BqDGNa8oEW5Xg7 
  * Please BE SPECIFIC so I can address concerns!
    * instead of 'more examples', try 'lecture X was challenging. more examples on Y'
* **Upcoming dates:**
  * University orientation (9/22)
  * MACSS Econ orientation (TBD)
  * MACSS orientation (9/23) + rest of week
  * Career services (9/25) (tentative: 10-7)
  * pub night (tbd)

---

# Learning objectives

* Define random variables
* Distinguish between discrete and continuous variables
* Identify discrete random variable distributions relevant to social science
* Review measures of central tendency and dispersion
* Define expected value and variance
* Define cumulative distribution functions (CDFs) for discrete random variables

---

# Random variable

* A random process or variable with a numerical outcome
* Random variable $X$ that is a function of the sample space
    * Number of incumbents who win
    * An indicator whether a country defaults on a loan (1 if a default, 0 otherwise)
    * Number of casualties in a war (rather than all possible outcomes)
    
    $$X:\text{Sample Space} \rightarrow \Re$$

---

# Treatment assignment

* $3$ units, flipping fair coin $\frac{1}{2}$ to assign each unit
* Assign to $T=$Treatment or $C=$control
* $X$ = Number of units received treatment
* Defining the function

    $$X  = \left \{ \begin{array} {ll}0  \text{  if  } (C, C, C)  \\1 \text{  if  } (T, C, C) \text{ or } (C, T, C) \text{ or } (C, C, T) \\2 \text{ if  }  (T, T, C) \text{ or } (T, C, T) \text{ or } (C, T, T) \\3 \text{ if } (T, T, T) \end{array} \right.$$
    
    $$\begin{aligned}X( (C, C, C) )  & = 0 \\X( (T, C, C)) & =  1 \\X((T, C, T)) & = 2 \\X((T, T, T)) & = 3 \end{aligned}$$

---

# Examples

* $X$ = Number of Calls into congressional office in some period $p$
    * $X(c) = c$
* Outcome of Election
    * Define $v$ as the proportion of vote the candidate receives
    * Define $X = 1$ if $v>0.50$
    * Define $X = 0$ if $v<0.50$
    * For example, if $v = 0.48$, then $X(v) = 0$
* An indicator whether a country defaults on a loan (1 if a default, 0 otherwise)
* Not all are experimental - most are observational

---

# Discrete random variables

* A random variable with a finite or countably infinite range is **discrete**
* A random variable taking uncountably many values, with a density rather than point masses, is **continuous**

---

# Probability mass functions

* Probability of the values that our value can take. 
* FOR DISCRETE VALUES ONLY!!
* Operator is the $\sum$ (sum) because we are ADDING PIECES
* Need a corresponding probability for each value (remember what we said re: probability ( $0 \leq p \leq 1$ ))


---
# Probability mass function (PMF)
Define these for our domain (x). NOTE: little 'x' is the value that is being taken on while big 'X' is the random variable itself. 

--

So, $P(X=x)$ refers to the probability that our random variable takes on the value of x. You can also see $P_X(X=x)$ and similar variants.

---

### Intuition

* $\Pr(C, T, C) = \Pr(C)\Pr(T)\Pr(C) = \frac{1}{2}*\frac{1}{2}*\frac{1}{2} = \frac{1}{8}$

    $$\begin{aligned}p(X = 0) & = \Pr(C, C, C) = \frac{1}{8}\\p(X = 1) & = \Pr(T, C, C) + \Pr(C, T, C) + \Pr(C, C, T) = \frac{3}{8} \\p(X = 2) & = \Pr(T, T, C)  + \Pr(T, C, T) + \Pr(C, T, T) = \frac{3}{8} \\p(X = 3) & = \Pr(T, T, T) = \frac{1}{8}\end{aligned}$$
    
* $p(X = a) = 0$, for all $a \notin {0, 1, 2, 3}$

---

# Probability mass function

```{r pmf}
pmf <- tibble(
  x = 0:3,
  y = c(1 / 8, 3 / 8, 3 / 8, 1 / 8)
)

pmf %>%
  ggplot(aes(x, y)) +
  geom_col() +
  labs(
    title = "Probability mass function",
    x = "Number treated",
    y = "Probability"
  )
```

---

# Probability mass function

* For a discrete random variable $X$

    $$p_X(x) = \Pr(X = x)$$
    
* Note that

    $$\sum_x p_{X}(x) = 1$$
  
  

--
  
    
* Can also add probabilities for smaller sets $S$ of possible values of $X \in S$

    $$\Pr(X \in S) = \sum_{x \in S} p_X (x)$$
    $$\Pr (X > 0) = \sum_{x=1}^2 p_X (x) = \frac{1}{2} + \frac{1}{4} = \frac{3}{4}$$

--

*remind you of anything* **::cough::** *integrals*
---

# Calculating our PMF!

To calculate the PMF of a random variable $X$

* For each possible value $x$ of $X$, collect all possible outcomes that give rise to the event $\{ X=x \}$
* Add their probabilities to obtain $p_X (x)$

---

# Topic model example

* Topics: distinct concepts (war in Ukraine, national debt, fire department grants)... Mathematically: PMF on words
* Set of words $\{ \text{ukraine, fire, department, soldier, troop, war, grant}\}$

--

### Topic 1

$\Pr(\text{ukraine}) = 0.3$; $\Pr(\text{fire}) = 0.0001$; $\Pr(\text{department}) = 0.0001$; $\Pr(\text{soldier}) = 0.2$; $\Pr(\text{troop}) = 0.2$; $\Pr(\text{war})=0.2997$; $\Pr(\text{grant})=0.0001$

--

### Topic 2

$\Pr(\text{ukraine}) = 0.0001$; $\Pr(\text{fire}) = 0.3$; $\Pr(\text{department}) = 0.2$; $\Pr(\text{soldier}) = 0.0001$; $\Pr(\text{troop}) = 0.0001$; $\Pr(\text{war})=0.0001$; $\Pr(\text{grant})=0.2997$

--

* Topic models: take a set of documents and estimate topics

---

# Cumulative mass function

* For a random variable $X$

    $$F(x) = \Pr(X \leq x)$$
  --
* Characterizes how probability **CUMULATES** as $X$ gets larger
* $F(x) \in [0,1]$
* $F(x)$ is non-decreasing

---

# Three person experiment

Consider the three person experiment: $\Pr(T) = \Pr(C) =  1/2$

--

What is $F(2)$?

$$\begin{aligned}F(2) & = \Pr(X = 0) + \Pr(X = 1) + \Pr(X = 2) \\& = \frac{1}{8} + \frac{3}{8} + \frac{3}{8} \\& = \frac{7}{8} \end{aligned}$$

--

What is $F(2) - F(1)$?

$$\begin{aligned} F(2)  - F(1) & = [\Pr(X = 0) + \Pr(X = 1) + \Pr(X = 2)]  \nonumber \\& \quad -[\Pr(X = 0) + \Pr(X = 1)] \\F(2) - F(1) & = \Pr(X = 2)\end{aligned}$$


---
class: middle, center

# PMFs and CMFs

PMF is to our original function as CMF is to an integral (ish)
---

# PMFs and CMFs

```{r cmf}
pmf %>%
  mutate(cumsum = cumsum(y),
         xend = x + 1) %>%
  ggplot(mapping = aes(x, cumsum)) +
  geom_segment(mapping = aes(xend = xend, yend = cumsum)) +
  ylim(0,1) +
  labs(
    title = "Cumulative mass function",
    x = "Number of treated",
    y = expression(F(x))
  )
```

* Similar to integration, but over a discrete set of values

---
class: middle, center
# Distributions and the functions we love

```{r}
knitr::include_graphics("https://encrypted-tbn0.gstatic.com/images?q=tbn:ANd9GcSbEyaYCtKvH69f_7DIShQ_Kc67Me-j5knH0A&usqp=CAU")
```

---
# Function & Distribution overview

So far, we've looked at how to add up values. However, the values we get depend upon the underlying function. There are a few well-known distributions that we'll cover as these will be increasingly useful as you move through your coursework. 

* Bernoulli
* Binomial
* Geometric
* Poisson 

--

This is a selection of distributions -- each has certain properties that can be useful for predicting outcomes and events but there are many, many more that may be useful to you as well. 


---
# Bernoulli

Suppose $X$ is a random variable, with $X \in \{0, 1\}$ and $\Pr(X = 1) = \pi$. Then we will say that $X$ is a **Bernoulli** random variable,

$$p_X(k)= \pi^{k} (1- \pi)^{1 - k}$$

for $k \in \{0,1\}$ and $p_X(k) = 0$ otherwise

--

$$Y \sim \text{Bernoulli}(\pi)$$

---

# Bernoulli

Suppose we flip a fair coin and $Y = 1$ if the outcome is Heads

$$
\begin{aligned}
Y & \sim \text{Bernoulli}(1/2) \nonumber \\
p_Y(1) & = (1/2)^{1} (1- 1/2)^{ 1- 1} = 1/2 \nonumber \\
p_Y(0) & = (1/2)^{0} (1- 1/2)^{1 - 0} = (1- 1/2) \nonumber 
\end{aligned}
$$

---

# Bernoulli PMF

```{r bernoulli, fig.cap = "Example Bernoulli probability mass functions"}
bernoulli_plot <- function(p) {
  tibble(
    x = 0:1,
    p = dbinom(x = x, size = 1, prob = p)
  ) %>%
    mutate(x = factor(x)) %>%
    ggplot(aes(x, p)) +
    geom_col() +
    labs(
      subtitle = bquote(pi == .(p)),
      x = expression(x),
      y = expression(p[X](k))
    )
}

bernoulli_plot(.5) +
  bernoulli_plot(.7) +
  bernoulli_plot(.3)
```

---

# Binomial

A model to count the number of successes across $N$ trials. 

--

Suppose $X$ is a random variable that counts the number of successes in $N$ independent and identically distributed Bernoulli trials.  Then $X$ is a **Binomial** random variable, 

--

$$p_X(k) = {N\choose{k}}\pi^{k} (1- \pi)^{N-k}$$
    
for $k \in \{0, 1, 2, \ldots, N\}$ and 0 otherwise

$$\binom N{k} = \frac{N!}{(N-k)! k!}$$

--
To say that a variable, $Y$, has or follows a binomial distribution we write: 
$$Y \sim \text{Binomial}(N, \pi)$$

---

# Binomial PMF: trials
--
```{r binomial}
binomial_plot <- function(n, p) {
  tibble(
    x = 0:n,
    p = dbinom(x = x, size = n, prob = p)
  ) %>%
    ggplot(aes(x, p)) +
    geom_col() +
    scale_x_continuous(breaks = scales::pretty_breaks()) +
    labs(
      subtitle = bquote(n == .(n) ~ p == .(p)),
      x = expression(x),
      y = expression(p[X](k))
    )
}

binomial_plot(10, .5) +
  binomial_plot(10, .3) +
  binomial_plot(100, .2) +
  binomial_plot(100, .8)
```

---

# Binomial

Recall our experiment example

* $\Pr(T) = \Pr(C) = 1/2$
* $Z =$ number of units assigned to treatment

--

$$
\begin{aligned}
Z & \sim  \text{Binomial}(3, 1/2)\\
p_Z(0) & = {3\choose0} (1/2)^0 (1- 1/2)^{3-0} = 1 * \frac{1}{8}\\
p_Z(1) & = {3\choose1} (1/2)^1 (1 - 1/2)^{2} = 3 * \frac{1}{8} \\
p_Z(2) & = {3\choose2} (1/2)^2 (1- 1/2)^1 = 3 * \frac{1}{8} \\
p_Z(3) & = {3\choose3} (1/2)^3 (1 - 1/2)^{0} = 1 * \frac{1}{8}
\end{aligned}
$$

---

# Geometric

* A model to **count the number of trials** of a Bernoulli **up to and including the FIRST success**
--

* Suppose $X$ is a random variable that counts the number of tosses needed for a head to come up the first time. Its PMF is 

    $$p_X(k) = (1 - p)^{k-1}p, \quad k = 1, 2, \ldots$$
    
--
We can then add up all these possibilities to ensure it satisfies our criteria for probability: 
$$
\begin{aligned}
\sum_{k=1}^{\infty} p_X(k) &= \sum_{k=1}^{\infty} (1 - p)^{k-1}p \\
&= p \sum_{k=1}^{\infty} (1 - p)^{k-1} \\&
= p * \frac{1}{1 - (1-p)} \\
&= 1
\end{aligned}
$$

---

# Geometric PMF

```{r geometric}
geometric_plot <- function(p) {
  tibble(
    x = 0:20,
    p = dgeom(x = x, prob = p)
  ) %>%
    mutate(x = x + 1) %>%
    ggplot(aes(x, p)) +
    geom_col() +
    scale_x_continuous(breaks = scales::pretty_breaks()) +
    labs(
      subtitle = bquote(p == .(p)),
      x = expression(x),
      y = expression(p[X](k))
    )
}

geometric_plot(.5) +
  geometric_plot(.7) +
  geometric_plot(.2)
```

---

# Poisson

* Often interested in counting number of events that occur
    * Number of wars started
    * Number of speeches made
    * Number of bribes offered
    * Number of people waiting for licenses
* Generally referred to as **event counts**

---

# Poisson

Suppose $X$ is a random variable that takes on values $X \in \{0, 1, 2, \ldots, \}$ and that $\Pr(X = k) = p_X(k)$ is,

$$p_X(k) = e^{-\lambda} \frac{\lambda^{k}}{k!}, \quad k = 0,1,2,\ldots$$
    
for $k \in \{0, 1, \ldots, \}$ and $0$ otherwise. (Here, $\lambda$ is the mean number of events per unit that we're interested in.)

--

$$X \sim \text{Poisson}(\lambda)$$

---

# Poisson PMF

```{r poisson}
poisson_plot <- function(lambda, max_n = 10) {
  tibble(
    x = 0:max_n,
    p = dpois(x = x, lambda = lambda)
  ) %>%
    ggplot(aes(x, p)) +
    geom_col() +
    scale_x_continuous(breaks = scales::pretty_breaks()) +
    labs(
      title = "Poisson PMF",
      subtitle = bquote(lambda == .(lambda)),
      x = expression(x),
      y = expression(p[X](k))
    )
}

poisson_plot(2) +
  poisson_plot(5.5) +
  poisson_plot(78, max_n = 150)
```

---

# Example

* Suppose the number of threats a president makes in a term is given by $X \sim \text{Poisson}(5)$
* What is the probability the president will make ten threats?

--

$$p_X(10) = e^{-5} \frac{5^{10}}{10!}$$

```{r poisson-president, fig.height = 5}
tibble(
  n = 0:50,
  prob = dpois(n, 5)
) %>%
  ggplot(aes(n, prob)) +
  geom_col() +
  labs(
    x = "Number of threats",
    y = "Probability"
  )
```

---
# Example: working backwards
Suppose we have an event which we believe follows a Poisson distribution. Perhaps this is the number of students who forget their calculator for the exam. If we observed that last year, about $5\%$ of the time all students remembered their calculator. How could we figure out the likelihood that exactly one student forgets theirs in the upcoming exam?

--

$$p_X(k) = e^{-\lambda} \frac{\lambda^{k}}{k!}$$
--
$$p_X(0) = e^{-\lambda} \frac{\lambda^{0}}{0!}=0.05= e^{-\lambda}$$
--
$$\log(0.05) = - \lambda \, \text{ therefore } \lambda \approx 3.0$$ -- $$p_X(1) = e^{-3}\frac{3^{1}}{1!} \approx 0.149$$
---
# Poisson plot

```{r poisson-student, fig.height = 5}
tibble(
  n = 0:50,
  prob = dpois(n, 3)
) %>%
  ggplot(aes(n, prob)) +
  geom_col() +
  labs(
    x = "Number of forgotten calculators",
    y = "Probability"
  )
```

---

# Approximating a binomial RV

The Poisson PMF with parameter $\lambda$ is a good approximation for a binomial PMF with parameters $n$ and $p$
    
$$e^{-\lambda} \frac{\lambda^{k}}{k!} \approx {n\choose{k}}p^{k} (1- p)^{n-k}, \quad \text{if } k \ll n$$

* Assumes
    * $\lambda = np$
    * $n$ is very large
    * $p$ is very small

---

# Approximating a binomial RV

* $n = 100$ and $p = 0.01$

--

#### Using the binomial PMF

$$\frac{100!}{95! 5!} * 0.01^5 (1 - 0.01)^{95} = 0.00290$$

--

#### Using the Poisson PMF

* $\lambda = np = 100 * 0.01 = 1$

$$e^{-1} \frac{1}{5!} = 0.00306$$
---
# Function recap:

* We're dealing here ONLY with **DISCRETE** functions
* Random variable takes on a limited number of values
* Functions offer a form for calculating probability
* Different assumptions for different types of functions

---
class: center, middle, inverse

# More fun with random variables
---

# Functions of random variables

Given a random variable $X$, you may wish to create a new random variable $Y$ using transformations of $X$

$$Y = g(X) = aX + b$$

$$g(X) = \log(X)$$

--

* If $Y = g(X)$ is a function of a random variable $X$, then $Y$ is also a random variable
* All outcomes in the sample space defined with a numerical value $x$ for $X$ also have a numerical value $y = g(x)$ for $Y$

---

# Expectation, mean, and variance

* PMF of a random variable $X$ provides probabilities of all possible values of $X$
* Often desirable to summarize this information into a single representative number
* **Expectation** of $X$

---

# Mean and Variance: Motivation

* Consider spinning a wheel of fortune many times
* At each spin, one of the numbers $m_1, m_2, \ldots, m_n$ comes up with corresponding probability $p_1, p_2, \ldots, p_n$, and this is your monetary reward from that spin
* What is the amount of money that you "expect" to get "per spin"?

--

* Spin $k$ times: $k_i$ and $m_i$
* $m_1 k_1 + m_2 k_2 + \ldots + m_n k_n$

    $$M = \frac{m_1 k_1 + m_2 k_2 + \ldots + m_n k_n}{k}$$

    $$\frac{k_i}{k} \approx p_i, i = 1, \ldots,n$$

    $$M = \frac{m_1 k_1 + m_2 k_2 + \ldots + m_n k_n}{k} \approx m_1p_1 + m_2p_2 + \ldots + m_np_n$$

---

# Expectation

Random variable $X$, with PMF $p_X$

$$\E[X] = \sum_{x:p(x)>0} x p(x)$$

---

# Example of expected value

Suppose $X$ is number of units assigned to treatment

$$X  = \left \{ \begin{array} {ll}0  \text{  if  } (C, C, C)  \\1 \text{  if  } (T, C, C) \text{ or } (C, T, C) \text{ or } (C, C, T) \\2 \text{ if  }  (T, T, C) \text{ or } (T, C, T) \text{ or } (C, T, T) \\3 \text{ if } (T, T, T) \end{array} \right.$$

What is $\E[X]$?

--

$$\begin{aligned}\E[X]  & = 0* \frac{1}{8} + 1 * \frac{3}{8} + 2 * \frac{3}{8} + 3 * \frac{1}{8} \\& = 1.5 \end{aligned}$$

--

* Weighted average
* Measure of central tendency

---

# A single person poll

Suppose that there is a group of $N$ people

* Suppose $M< N$ people approve of the president's performance
* $N - M$ disapprove of their performance

--

Draw one person $i$, with $\Pr(\text{Draw } i ) = \frac{1}{N}$

$$X  = \left \{ \begin{array} {ll}	1  \text{  if  person Approves}  \\    0 \text{  if  Disapproves}   \\\end{array} \right.$$

--

What is $\E[X]$?

--

$$\begin{aligned}\E[X] & = 1 * \Pr(\text{Approve})  + 0 * \Pr(\text{Disapprove}) \\ & = 1 * \frac{M}{N} \\ & = \frac{M}{N}  \end{aligned}$$

---

# Indicator variables and probabilities

Suppose $A$ is an event. Define random variable $I$ such that $I= 1$ if an outcome in $A$ occurs and $I =0$ if an outcome in $A^{c}$ occurs.

Then,

$$\E[I] = \Pr(A)$$

--


$$\begin{aligned}\E[I]  & =  1 * \Pr(A) + 0 * \Pr(A^{c}) \\ & = \Pr(A) \end{aligned}$$

---

# Moments

* 1st moment: $\E[X^1] = \E[X]$
* 2nd moment: $\E[X^2]$
* $n$th moment: $\E[X^n]$

---

# Variance

* Expected value is a measure of **central tendency**
* What about **spread** or **dispersion**?

--

For each value, measure distance from the center. Recall that $E[X]$ is a SET VALUE.

$$d(x, E[X])^{2} = (x - E[X])^2$$

--

$$\begin{aligned} \E[(X - \E[X])^2] & = \sum_{x:p_X(x)>0} (x  - \E[X])^2p_X(x)\end{aligned}$$
$$=\sum_{x:p_X(x)>0} p_X(x)[\left(x^2 \right)- 2 \E[X] \left(x  \right) + \E[X]^2  ]$$
$$=\sum_{x:p_X(x)>0} [\left(x^2 p_X(x)\right)- 2 \E[X] \left(x p_X(x)\right) + \E[X]^2  p_X(x)]$$

---

Weighted average of these distances
    
$$\sum_{x:p_X(x)>0} [\left(x^2 p_X(x)\right)- 2 \E[X] \left(x p_X(x)\right) + \E[X]^2  p_X(x)]$$
--

$$\sum_{x:p_X(x)>0} \left(x^2 p_X(x)\right)- 2 \E[X]\sum_{x:p_X(x)>0} \left(x p_X(x)\right) + \E[X]^2\sum_{x:p_X(x)>0} p_X(x)$$
--

$$=  \E[X^2] - 2\E[X]^2 + \E[X]^2 \\ = \E[X^2] - \E[X]^2 \\= \text{Var}(X)$$

---

# Variance

Expected value of the random variable $(X - \E[X])^2$

$$\begin{aligned}\Var(X) &= \E[(X - \E[X])^2] \\&= \E[X^2] - \E[X]^2\end{aligned}$$
    
* Measure of **dispersion** of $X$ around its mean
* Standard deviation of $X$ - $\sigma_X = \sqrt{\Var(X)}$

---

# Calculating variance of a random variable

* Generate the PMF of the random variable $(X - \E[X])^2$
* Calculate the expectation of this function

--

$$\begin{align}\Var(X) &= \E[(X - \E[X])^2] \\\Var(X) &= \E[X^2] - \E[X]^2\end{align}$$

---

## Expected value rule for functions of random variables

Let $X$ be a random variable with PMF $p_X$, and let $g(X)$ be a function of $X$. Then, the expected value of the random variable $g(X)$ is given by

$$\E[g(X)] = \sum_{x} g(x) p_X(x)$$


    
---

# Bernoulli variable

* Suppose $Y \sim \text{Bernoulli}(\pi)$

    $$\begin{aligned} \E[Y] & = 1 * \Pr(Y = 1) + 0 * \Pr(Y = 0) \nonumber \\& = \pi + 0 (1 - \pi) \nonumber  = \pi \\\Var(Y) & = \E[Y^2] - \E[Y]^2 \nonumber  \\\E[Y^2] & = 1^{2} \Pr(Y = 1) + 0^{2} \Pr(Y = 0) \nonumber \\& = \pi \nonumber \\ \Var(Y) & = \pi - \pi^{2} \nonumber \\& = \pi(1 - \pi ) \nonumber\end{aligned}$$

--

* $\E[Y] = \pi$
* $\Var(Y) = \pi(1- \pi)$
* What is the maximum variance?

--

$Var(Y) = \pi(1- \pi)$ Derivative and set to zero: $1-2\pi = 0 \rightarrow \pi = 0.5$
So: max variance at $0.5(1 - 0.5 ) = 0.25$

---

# Binomial

$$Z = \sum_{i=1}^N Y_{i} \text{ where } Y_{i} \sim \text{Bernoulli}(\pi)$$

--

$$\begin{aligned}\E[Z] & = \E[Y_{1} + Y_{2} + Y_3 + \ldots + Y_N ] \\& = \sum_{i=1}^N \E[Y_{i} ] \\& = N \pi \\\Var(Z) & = \sum_{i=1}^N \Var(Y_{i}) \\& = N \pi (1-\pi)\end{aligned}$$

---

# Decision making using expected values

* Optimizes the choice between several candidate decisions that result in random rewards
* View the expected reward of a decision as its average payoff over a large number of trials
* Choose a decision with maximum expected reward

---

# Example: going to war

* Suppose country $1$ is engaged in a conflict and can either win or lose
* Define $Y = 1$ if the country wins and $Y = 0$ otherwise

    $$Y \sim \text{Bernoulli}(\pi)$$

--

* Suppose country $1$ is deciding whether to fight a war
* Engaging in the war will cost the country $c$
* If they win, country $1$ receives $B$
* What is $1$'s expected utility from fighting a war?

--

$$\begin{aligned}\E[U(\text{war})] & = U(\text{war} | \text{win}) * \Pr(\text{win}) + U(\text{war} | \text{lose}) * \Pr(\text{lose}) \\& = (B - c) \Pr(Y = 1) + (- c) \Pr(Y = 0 ) \\& = B * \Pr(Y = 1)  - c(\Pr(Y = 1)  + \Pr(Y = 0)) \\& = B * \pi - c \end{aligned}$$

---

# Cumulative mass function

* Defines the cumulative probability $F_X(x)$ up to the value of $x$
* For a discrete random variable $X$

    $$F_X(x) = \Pr (X \leq x) = \sum_{k \leq x} p_X(k)$$

* All random variables with a PMF have a CMF

---

# Cumulative mass function

* $F_X$ is monotonically non-decreasing - if $x \leq y$, then $F_X(x) \leq F_X(y)$
* $F_X(x)$ tends to $0$ as $x \rightarrow -\infty$, and to $1$ as $x \rightarrow \infty$
* $F_X(x)$ is a piecewise constant function of $x$
* If $X$ is discrete and takes integer values, the PMF and the CMF can be obtained from each other by summing or differencing:

    $$F_X(k) = \sum_{i = -\infty}^k p_X(i),$$
    $$p_X(k) = \Pr (X \leq k) - \Pr (X \leq k-1) = F_X(k) - F_X(k-1)$$
    
    for all integers $k$
    
---

# Bernoulli CMF

```{r bernoulli-cmf, fig.cap = "Bernoulli random variable"}
bernoulli_cmf_plot <- function(p) {
  df <- tibble(
    x = 0:1,
    pmf = dbinom(x = x, size = 1, prob = p),
    cmf = pbinom(q = x, size = 1, prob = p)
  )

  ggplot(df, aes(x, pmf)) +
    geom_col() +
    ylim(0,1) + scale_x_continuous(breaks = 0:1) +
    labs(
      subtitle = bquote(pi == .(p)),
      x = expression(x),
      y = expression(p[X](k))
    ) +
    ggplot(df, aes(x, cmf)) + ylim(0,1) +
    scale_x_continuous(breaks = 0:1) +
    geom_col() +
    labs(
      subtitle = bquote(pi == .(p)),
      x = expression(x),
      y = expression(F[X](k))
    )
}

bernoulli_cmf_plot(.7)
```

---

# Binomial CMF

```{r binomial-cmf, fig.cap = "Binomial random variable"}
binomial_cmf_plot <- function(n, p) {
  df <- tibble(
    x = 0:n,
    pmf = dbinom(x = x, size = n, prob = p),
    cmf = pbinom(q = x, size = n, prob = p)
  )

  ggplot(df, aes(x, pmf)) +
    geom_col() + ylim(0,1) +
    scale_x_continuous(breaks = scales::pretty_breaks()) +
    labs(
      subtitle = bquote(n == .(n) ~ p == .(p)),
      x = expression(x),
      y = expression(p[X](k))
    ) +
    ggplot(df, aes(x, cmf)) +
    geom_col() +
    scale_x_continuous(breaks = scales::pretty_breaks()) +
    labs(
      subtitle = bquote(n == .(n) ~ p == .(p)),
      x = expression(x),
      y = expression(F[X](k))
    )
}

binomial_cmf_plot(10, .5)
```

---

# Binomial CMF

```{r binomial-cmf-2, dependson = "binomial-cmf"}
binomial_cmf_plot(10, .3)
```

---

# Binomial CMF

```{r binomial-cmf-3, dependson = "binomial-cmf"}
binomial_cmf_plot(100, .2)
```

---

# Binomial CMF

```{r binomial-cmf-4, dependson = "binomial-cmf"}
binomial_cmf_plot(100, .8)
```

---

# Geometric CMF

```{r geometric-cmf, fig.cap = "Geometric random variable"}
geometric_cmf_plot <- function(p) {
  df <- tibble(
    x = 0:20,
    pmf = dgeom(x = x, prob = p),
    cmf = pgeom(q = x, prob = p)
  ) %>%
    mutate(x = x + 1)

  ggplot(df, aes(x, pmf)) +
    geom_col() + ylim(0,1) +
    scale_x_continuous(breaks = scales::pretty_breaks()) +
    labs(
      subtitle = bquote(p == .(p)),
      x = expression(x),
      y = expression(p[X](k))
    ) +
    ggplot(df, aes(x, cmf)) +
    geom_col() +
    scale_x_continuous(breaks = scales::pretty_breaks()) +
    labs(
      subtitle = bquote(p == .(p)),
      x = expression(x),
      y = expression(F[X](k))
    )
}

geometric_cmf_plot(.5)
```

---

# Poisson CMF

```{r poisson-cmf, fig.cap = "Poisson random variable"}
poisson_cmf_plot <- function(lambda, max_n = 10) {
  df <- tibble(
    x = 0:max_n,
    pmf = dpois(x = x, lambda = lambda),
    cmf = ppois(q = x, lambda = lambda)
  )

  ggplot(df, aes(x, pmf)) +
    geom_col() + ylim(0,1) +
    scale_x_continuous(breaks = scales::pretty_breaks()) +
    labs(
      subtitle = bquote(lambda == .(lambda)),
      x = expression(x),
      y = expression(p[X](k))
    ) +
    ggplot(df, aes(x, cmf)) +
    geom_col() +
    scale_x_continuous(breaks = scales::pretty_breaks()) +
    labs(
      subtitle = bquote(lambda == .(lambda)),
      x = expression(x),
      y = expression(F[X](k))
    )
}

poisson_cmf_plot(2)
```

---

# RECAP:

* Random variables: $P(X=x)$
* DISCRETE VARIABLES: **different from continuous**
* PMF: probability mass function, Probability for a particular value
* CDF: cumulative distribution function, ADD UP VALUES (akin to a sum)
* **Distributions**: 
  * Bernoulli: 
      * pmf: $1-\pi$ for $k = 0$; $\pi$ for $k = 1$
      * mean $p$, var $p(1-p)$
  * Binomial: 
      * pmf: $p_X(k) = {N\choose{k}}\pi^{k} (1- \pi)^{N-k}$
      * mean $np$, var $np(1-p)$ (also written $npq$)
  * Geometric: 
      * pmf: $p_X(k) = (1 - p)^{k-1}p, \quad k = 1, 2, \ldots$
      * mean $\frac{1}{p}$ , var $\frac{(1-p)}{p^2}$
  * Poisson: 
      * pmf: $p_X(k) = e^{-\lambda} \frac{\lambda^{k}}{k!}, \quad k = 0,1,2,\ldots$
      * mean $\lambda$, var $\lambda$
      


---
class: center, middle, inverse
## Let’s get to know each other a bit more – Name, pronouns, subfield/research area, where you are currently, something fun/interesting about you and/or your hobbies
