---
title: "General random variables"
author: "[MACS 33000](https://jmclip.github.io/MACSS_math_camp/) <br /> University of Chicago"
output:
  xaringan::moon_reader:
    css: ["xaringan-themer.css", "slides-menu.css"]
    math: katex
    includes:
      after_body: [slides-menu.js]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false 
---

```{r child = here::here("_common-slides.Rmd"), cache = FALSE}

```

```{r packages, include=FALSE}
library(tidyverse)
library(patchwork)
library(here)
library(gganimate)
```

$$\newcommand{\E}{\mathrm{E}} \newcommand{\Var}{\mathrm{Var}} \newcommand{\Cov}{\mathrm{Cov}}$$
# Misc

* Assigned rooms for final exam (two locations -- other is on other side of the quad)
* Calculators: 5 left! 
* Will be drafting exam questions today and tomorrow -- submit your questions!



---
class: center, middle, inverse

# Survey Results



---
# Survey:

Overall course is going well. Wide range of thoughts on pacing from "it's great!" to "way too fast!"

* **Immediate changes**: more mindful of boardwork, no blue/purple chalk (sorry!)
* **Later changes**: revisit pretest idea (excellent critique for why not, however!), final exam timing, notation explanation
* **Eventual changes**: more practice questions with worked solutions
* **Suggestions I'm thinking about**: case studies for context, other ideas for pacing, different feelings about first couple of days


---

# Learning objectives

* Define a continuous random variable
* Identify continuous random variable distributions relevant to social science
* Define expected value and variance of continuous random variables
* Relate continuous random variables to discrete random variables
* Define cumulative distribution functions (CDFs) for continuous random variables and compare to discrete random variables
* Estimate probability of events using probability density functions (PDFs) and cumulative distribution functions (CDFs)

---

# Continuous random variables

* Random variables that are not discrete
    * Approval ratings
    * GDP
    * Wait time between wars: $X(t) = t$ for all $t$
    * Proportion of vote received: $X(v) = v$ for all $v$
* Many analogues to discrete probability distributions
* We need calculus to answer questions about probability

---

# Probability density function

```{r pdf, fig.height = 6}
norm_pdf <- tibble(
  x = seq(-3, 3, by = .05),
  y = dnorm(x = x, mean = 1, sd = 2),
  fill = x >= .5 & x <= 2
)

ggplot(norm_pdf) +
  geom_line(aes(x, y)) +
  geom_ribbon(
    data = filter(norm_pdf, fill),
    aes(
      x = x,
      ymin = 0, ymax = y
    ),
    fill = "#0B6884"
  ) +
  labs(
    x = expression(x),
    y = expression(f(x))
  )
```

--

What is the area under the curve under $f(x)$ between $.5$ and $2$?

--

$$\int_{1/2}^{2} f(x)\,dx = F(2) - F(1/2)$$

---

# Continuous random variable

$X$ is a **continuous random variable** if there exists a nonnegative function $f_X$ defined for all $x \in \Re$ having the property for any (measurable) set of real numbers $B$,

$$\Pr(X \in B) = \int_{B} f_X(x)\,dx$$

---

# Continuous random variable

The probability that the value of $X$ falls within an interval is

$$\Pr (a \leq X \leq b) = \int_a^b f_X(x) \,dx$$

--
Compare discrete and continuous variables. Suppose we have a probability function $f(x)= 1/3$ where $0 < x \leq 3$. What is $p(X=x)$?

--

* For any single value $a$, $\Pr (X = a) = \int_a^a f_X(x) \,dx = 0$

--

### Requirements to be a PDF

* Non-negative
* Normalization property

    $$\int_{-\infty}^{\infty} f_X(x) \,dx = \Pr (-\infty \leq X \leq \infty) = 1$$

---

# Uniform random variable

$$X \sim \text{Uniform}(0,1)$$

$$f_X(x) = \left\{    \begin{array}{ll}        c & \quad \text{if } 0 \leq x \leq 1 \\        0 & \quad \text{otherwise}    \end{array}\right.$$

--

$$1 = \int_{-\infty}^{\infty} f_X(x)\,dx = \int_0^1 c \,dx = c \int_0^1 \,dx = c$$

---

# Uniform random variable: plot

```{r unif}
tibble(
  x = seq(-.5, 1.5, by = .01),
  y = dunif(x)
) %>%
  ggplot(aes(x, y)) +
  geom_step() +
  labs(
    x = expression(x),
    y = expression(f[X](x))
  )
```

---

# Uniform random variable

$$\begin{aligned}
\Pr(X \in [0.2, 0.5]) & = \int_{0.2}^{0.5} 1 \,dx \\
& = x\big|^{0.5}_{0.2} = 0.5  - 0.2  \\
& = 0.3 
\end{aligned}$$

--

$$\begin{aligned}
\Pr(X \in [0.5, 0.5]) & = \int_{0.5}^{0.5} 1\,dx \\
& = X|^{0.5}_{0.5} = 0.5 - 0.5 \\
& = 0 
\end{aligned}$$

--

$$\begin{aligned}
\Pr(X \in \{[0, 0.2]\cup[0.5, 1]\}) & = \int_{0}^{0.2} 1\,dx + \int_{0.5}^{1} 1\,dx \\
 & = x\big|_{0}^{0.2} + x\big|_{0.5}^{1}  = 0.2 - 0 + 1 - 0.5 \\
 & = 0.7 
\end{aligned}$$

---

# Uniform random variable

$$f_X(x) = \left\{    \begin{array}{ll}        \frac{1}{b-a} & \quad \text{if } a \leq x \leq b \\        0 & \quad \text{otherwise}    \end{array}\right.$$

---

# Continuous random variables

* $\Pr(X = a) = 0$
* $\Pr(X \in (-\infty, \infty) ) = 1$ 
* If $F$ is antiderivative of $f$, then $\Pr(X \in [c,d]) = F(d) - F(c)$

---
class: middle, center, inverse

# Having a moment

---
# Moments: what they are and what they tell us
Moments help us understand the shape of the distribution

1. Center (mean): $E[X]$ -- the 1st raw moment
1. Width (variance): $E[X^2] - (E[X])^2$ -- built from the 2nd moment
1. Lopsidedness (skew): $E\left[\left(\frac{X-\mu}{\sigma}\right)^3\right]$ -- the 3rd standardized moment
1. Tail-y-ness (kurtosis): $E\left[\left(\frac{X-\mu}{\sigma}\right)^4\right]$ -- the 4th standardized moment

---
# Great expectations: first moment

--

* Summarizing a distribution
* What we can expect FROM THE DISTRIBUTION re: center, ON AVERAGE
* This will give us our baseline expectation for what we *should* be observing *if* our assumptions are true

---

# Expectation

$$\E[X] = \int_{-\infty}^{\infty} x f_X(x) \,dx$$

* Integration instead of summation

--

* Expected value rule

    $$\E[g(X)] = \int_{-\infty}^{\infty} g(x) f_X(x) \,dx$$

--

* $n$th moment = $\E[X^n]$

--

* $\Var(X) = E[(X-\mu)^2]=E[(X - \E[X])^2] = \E[X^2] - (\E[X])^2$

---

# Uniform random variable: Mean
Start with our pdf: $f(x) = \frac{1}{b-a}$ (defined over a to b).

--

$$\E[X] = \int_{-\infty}^{\infty} x f_X(x) \,dx$$
--
$$= \int_a^b x * \frac{1}{b-a} \,dx$$
--
$$= \frac{1}{b-a} * \frac{1}{2}x^2 \Big|_a^b$$
--
$$= \frac{1}{b-a} * \frac{b^2 - a^2}{2}= \frac{a+b}{2}$$
---
# Uniform random variable: Mean
Start with our pdf: $f(x) = \frac{1}{b-a}$ (defined over a to b). $$\mu = \frac{a+b}{2}$$

SO WHAT?????

--

This is what we can anticipate as the mean of our distribution. 
--
On average. 
--
Over time. 
--
In the long run. 

---

# Uniform random variable: Variance
Start with our pdf: $f(x) = \frac{1}{b-a}$ (defined over a to b). We need to recall two things: the formula for $V(X) \text{ and } E(X)$ 

$$\E[X^2] = \int_a^b x^2 * f(x) \,dx$$
--
$$\int_a^b x^2 * \frac{1}{b-a} \,dx$$
--
$$= \frac{1}{b-a} \int_a^b x^2 \,dx = \frac{1}{b-a} * \frac{1}{3}x^3 \Big|_a^b$$
--

$$= \frac{b^3 - a^3}{3(b-a)} = \frac{a^2 + ab + b^2}{3}$$

---
# Uniform random variable: Variance Cont'd

So, we can just plug these values into our formula:

$$\Var(X) = \E[X^2] - (\E[X])^2 = \frac{a^2 + ab + b^2}{3} - \left( \frac{a+b}{2} \right)^2 = \frac{(b-a)^2}{12}$$


---

# Exponential random variable

$$f_X(x) = \left\{    \begin{array}{ll}        \lambda e^{-\lambda x} & \quad \text{if } x \geq 0 \\        0 & \quad \text{otherwise}    \end{array}\right.$$

* $\lambda > 0$

--

* NOTE: you will sometimes see this with $\lambda$ replaced with $\frac{1}{\beta}$. 

--

* SUPER FUN NOTE: You can find its expected value yourself using integration by parts!!

---

# Exponential random variable

```{r exp-rv, fig.height = 6}
function_plot +
  stat_function(mapping = aes(color = "Exponential(1)"), fun = dexp, args = list(rate = 1), linewidth = 1.5) +
  stat_function(mapping = aes(color = "Exponential(3)"), fun = dexp, args = list(rate = 3), linewidth = 1.5) +
  stat_function(mapping = aes(color = "Exponential(5)"), fun = dexp, args = list(rate = 5), linewidth = 1.5) +
  scale_color_brewer(type = "qual") +
  xlim(0, 5) +
  labs(y = expression(f[X](x)),
       color = NULL)
```

--

$$\E[X] = \frac{1}{\lambda}, \quad \Var(X) = \frac{1}{\lambda^2}$$

---

# Cumulative distribution function

For a continuous random variable $X$ define its **cumulative distribution function** (CDF) $F_X(x)$ as, 

$$F_X(x) = \Pr(X \leq x) = \int_{-\infty} ^{x} f_X(t) \,dt$$

---

# Uniform distribution

Suppose $X \sim \text{Uniform}(0,1)$

$$\begin{aligned} F_X(x) & = \Pr(X\leq x)  \\& = 0 \text{, if }x< 0 \\ & = 1 \text{, if }x >1 \\ & = x \text{, if } x \in [0,1]\end{aligned}$$

--

```{r unif-rv, fig.height = 5}
unif_pdf <- function(a, b) {
  tibble(
    x = seq(a - 1, b + 1, by = .01),
    pdf = dunif(x = x, min = a, max = b)
  ) %>%
    ggplot(aes(x, pdf)) +
    geom_line() +
    labs(
      title = "Uniform PDF",
      subtitle = bquote(a == .(a) ~ b == .(b)),
      x = expression(x),
      y = expression(f[X](x))
    )
}

unif_cdf <- function(a, b) {
  tibble(
    x = seq(a - 1, b + 1, by = .01),
    cdf = punif(q = x, min = a, max = b)
  ) %>%
    ggplot(aes(x, cdf)) +
    geom_line() +
    labs(
      title = "Uniform CDF",
      subtitle = bquote(a == .(a) ~ b == .(b)),
      x = expression(x),
      y = expression(F[X](x))
    )
}

unif_pdf(0, 1) +
  unif_cdf(0, 1)
```

---

# Properties of CDFs

* $F_X$ is monotonically nondecreasing
* $F_X(x)$ tends to $0$ as $x \rightarrow -\infty$, and to $1$ as $x \rightarrow \infty$
* $F_X(x)$ is a continuous function of $x$
* If $X$ is continuous, the PDF and CDF can be obtained from each other by integration or differentiation

    $$F_X(x) = \int_{-\infty}^x f_X(t) \,dt, \quad f_X(x) = \frac{dF_X}{dx} (x)$$

---

# Normal distribution

Suppose $X$ is a random variable with $X \in \Re$

$$f(x) = \frac{1}{\sqrt{2\pi \sigma^2}}\exp\left(-\frac{(x - \mu)^2}{2\sigma^2}\right)$$

--

$$X \sim \text{Normal}(\mu, \sigma^2)$$

---

# Normal distribution

```{r norm-rv, fig.height = 6}
norm_pdf <- function(mu, sigma) {
  tibble(
    x = seq(-5, 5, by = .01),
    pdf = dnorm(x = x, mean = mu, sd = sigma)
  ) %>%
    ggplot(aes(x, pdf)) +
    geom_line() +
    labs(
      title = "Normal PDF",
      subtitle = bquote(mu == .(mu) ~ sigma^2 == .(sigma^2)),
      x = expression(x),
      y = expression(f[X](x))
    )
}

norm_cdf <- function(mu, sigma) {
  tibble(
    x = seq(-5, 5, by = .01),
    cdf = pnorm(q = x, mean = mu, sd = sigma)
  ) %>%
    ggplot(aes(x, cdf)) +
    geom_line() +
    labs(
      title = "Normal CDF",
      subtitle = bquote(mu == .(mu) ~ sigma^2 == .(sigma^2)),
      x = expression(x),
      y = expression(f[X](x))
    )
}

norm_pdf(0, 1) +
  norm_cdf(0, 1)
```

---

# Expected value/variance

* $Z$ is a standard normal distribution if

    $$Z \sim \text{Normal}(0,1)$$

* CDF of $Z$

    $$F_{Z}(x) = \frac{1}{\sqrt{2\pi} }\int_{-\infty}^{x} \exp(-z^2/2) \,dz$$

---

# Expected value/variance

Suppose $Z \sim \text{Normal}(0,1)$

* $Y = 2Z + 6$
* $Y \sim \text{Normal}(6, 4)$

    ```{r z-norm, fig.height = 6}
tibble(
  x = seq(-5, 15, by = .05),
  Z = dnorm(x),
  Y = dnorm(x, mean = 6, sd = 2)
) %>%
  gather(norm, val, -x) %>%
  ggplot(aes(x, val, color = norm)) +
  geom_line(linewidth = 1.5) +
  scale_color_brewer(type = "qual") +
  labs(
    x = NULL,
    y = "Density height",
    color = NULL
  ) +
  theme(legend.position = "bottom")
    ```

---

# Expected value/variance

If $Z \sim N(0,1)$, then $Y = aZ + b$ is

$$Y \sim \text{Normal} (b, a^2)$$

--

Assume we know

$$
\begin{aligned}
\E[Z]  & = 0 \\
\Var(Z) & = 1 
\end{aligned}
$$

--

For $Y \sim \text{Normal}(\mu, \sigma^2)$

$$\begin{aligned} \E[Y] & = \E[\sigma Z + \mu] \\& = \sigma \E[Z] + \mu \\& = \mu \\\Var(Y) & = \Var(\sigma Z + \mu) \\ & = \E[(\sigma Z + \mu)^2] - \left(\E[\sigma Z + \mu]\right)^2\\&= \sigma^2\E[Z^2] + 2\sigma\mu\E[Z] + \mu^2 - \mu^2 \\ & = \sigma^2 \Var(Z) \\ & = \sigma^2 \end{aligned}$$

---

# Standard normal distribution

If $X$ is a normal random variable with mean $\mu$ and variance $\sigma^2$, and if $a \neq 0, b$ are scalars, then the random variable

$$Y = aX + b$$

is also normal, with mean and variance

$$\E[Y] = a\mu + b, \quad \Var(Y) = a^2 \sigma^2$$

--

### Why rely on the standard normal distribution

* Normal distribution is commonly used in statistical analysis
* Ease of standardization
* Saves time on the calculus

---

# Support for President 

Suppose we are interested in modeling presidential approval

* $Y$: proportion of population who "approves of the job the president is doing"
* Individual responses are independent and identically distributed
* Average of those individual binary responses
* $N\rightarrow \infty$
* By Central Limit Theorem, $Y$ is normally distributed

$$
\begin{aligned}
Y & \sim \text{Normal}(\mu, \sigma^2) \\
f_Y(y) & = \frac{1}{\sqrt{2\pi \sigma^2}} \exp\left(-\frac{(y-\mu)^2}{2\sigma^2} \right)
\end{aligned}
$$

---

# Central limit theorem

```{r clt, out.width="90%", fig.height = 6}
# set.seed(124)
# y <- rnorm(7000, mean = .39, sd = sqrt(.0025))
# 
# clt_plot <- seq(from = 2, to = length(y), length.out = 100) %>%
#   round() %>%
#   map_dfr(~ y[1:.x] %>%
#             enframe(name = "y") %>%
#             mutate(nobs = .x)) %>%
#   ggplot(mapping = aes(x = value)) +
#   geom_density() +
#   stat_function(
#     fun = dnorm, n = 101, args = list(mean = .39, sd = sqrt(.0025)),
#     color = "#0B6884"
#   ) +
#   labs(
#     title = "Number of observations = {closest_state}",
#     x = "Mean",
#     y = "Density"
#   ) +
#   transition_states(states = nobs)
# 
# animate(clt_plot, nframes = 600)

set.seed(124)

# Each replication draws n Bernoulli(0.44) responses and keeps their
# STANDARDIZED mean, so the reference curve is N(0,1) at every n.
p <- 0.44
clt_plot <- map_dfr(c(2, 5, 10, 30, 100, 500), function(n) {
  m <- replicate(2000, mean(rbinom(n, size = 1, prob = p)))
  tibble(nobs = n, value = (m - p) / sqrt(p * (1 - p) / n))
}) %>%
  ggplot(mapping = aes(x = value)) +
  geom_density() +
  stat_function(fun = dnorm, n = 101, color = "#0B6884") +
  xlim(-4, 4) +
  labs(
    title = "Sample size = {closest_state}",
    x = "Standardized sample mean",
    y = "Density"
  ) +
  transition_states(states = nobs)
```

---

# Back to polling: presidential approval

* Suppose $\mu = 0.44$ and $\sigma^2 = 0.0025$
* $\Pr(Y\geq 0.49)$ 

--

$$\begin{aligned} \Pr(Y \geq 0.49)  & = 1 - \Pr(Y \leq 0.49 ) \\& = 1 - \Pr\left(Z \leq \frac{0.49-0.44 }{0.05} \right) \\& = 1 - \frac{1}{\sqrt{2\pi} } \int_{-\infty}^{1} \exp(-z^2/2) \,dz \\& = 1 - F_{Z} (1 ) \\& = 0.159\end{aligned}$$

---
# Gamma distribution: def

We use the gamma distribution when we're looking at something that takes on positive values and is skewed (long right tail). 

Example: waiting time between events

---

# Gamma distribution: formula

Suppose $\alpha>0$. Define $\Gamma(\alpha)$ as

$$
\begin{aligned}
\Gamma(\alpha) &= \int_{0}^{\infty} y^{\alpha- 1} e^{-y} \,dy \\
&= (\alpha- 1)! \, \forall \alpha \in \{1, 2, 3, \ldots\}
\end{aligned}
$$

--

* $\Gamma(\frac{1}{2}) = \sqrt{\pi}$

<!-- --- -->

<!-- # Gamma distribution: simplified formula -->

<!-- Suppose we have $\Gamma(\alpha)$ -->

<!-- $$\begin{aligned} -->
<!-- \frac{\Gamma(\alpha)}{\Gamma(\alpha)} & = \frac{\int_{0}^{\infty} y^{\alpha-1} e^{-y} dy}{\Gamma(\alpha)} \\ -->
<!-- 1 & = \int_{0}^{\infty} \frac{1}{\Gamma(\alpha)} y^{\alpha-1} e^{-y} \,dy  -->
<!-- \end{aligned}$$ -->

<!-- -- -->

<!-- Set $X = Y/\beta$ -->

<!-- -- -->

<!-- $$\begin{aligned} -->
<!-- F(x) = \Pr(X \leq x) & = \Pr(Y/\beta \leq x ) \\ -->
<!-- & = \Pr(Y \leq x \beta ) \\ -->
<!-- & = F_{Y} (x \beta) \\ -->
<!-- \frac{\partial F_{Y} (x \beta) }{\partial x} & = f_{Y} (x \beta) \beta -->
<!-- \end{aligned}$$ -->

<!-- -- -->

<!-- $$f(x|\alpha, \beta) = \frac{\beta^{\alpha}}{\Gamma(\alpha)} x^{\alpha - 1} e^{-x\beta}$$ -->

---

# Gamma distribution

Suppose $X$ is a continuous random variable, with $X \geq 0$. $X$ is a Gamma random variable if

$$f(x|\alpha, \beta) = \frac{\beta^{\alpha}}{\Gamma(\alpha)} x^{\alpha - 1} e^{-x\beta}$$

if $x\geq 0$ and $0$ otherwise

--

$$X \sim \text{Gamma}(\alpha, \beta)$$

---

# Gamma distribution

Suppose $X \sim \text{Gamma}(\alpha, \beta)$

$$\begin{aligned}
\E[X] & = \frac{\alpha}{\beta} \\
\Var(X) & = \frac{\alpha}{\beta^2}
\end{aligned}$$

--

Suppose $\alpha = 1$ and $\beta = \lambda$. If

$$\begin{aligned}
X & \sim \text{Gamma}(1, \lambda) \\
f(x|1, \lambda ) & = \lambda e^{- x \lambda}
\end{aligned}$$

$$X \sim \text{Exponential}(\lambda)$$

---

# Properties of Gamma distributions

Suppose we have a sequence of independent random variables, with

$$X_{i} \sim \text{Gamma}(\alpha_{i}, \beta)$$

--

Then

$$Y = \sum_{i=1}^{N} X_{i}$$

$$Y \sim \text{Gamma}(\sum_{i=1}^{N} \alpha_{i} , \beta)$$

---

# Gamma distribution v Exponential: plot

```{r gamma-3, fig.height = 6}
function_plot +
  stat_function(
    mapping = aes(color = "Exponential(1)"),
    fun = dexp, args = list(rate = 1), n = 501, linewidth = 1.5
  ) +
  stat_function(
    mapping = aes(color = "Exponential(5)"),
    fun = dexp, args = list(rate = 5), n = 501, linewidth = 1.5
  ) +
  stat_function(
    mapping = aes(color = "Gamma(1, 5.5)"),
    fun = dgamma, args = list(shape = 1, rate = 5.5), n = 501, linewidth = 1.5
  ) +
  stat_function(
    mapping = aes(color = "Gamma(3, 2)"),
    fun = dgamma, args = list(shape = 3, rate = 2), n = 501, linewidth = 1.5
  ) +
  scale_color_brewer(type = "div") +
  xlim(0, 10) +
  labs(
    y = expression(f(x)),
    color = NULL
  ) +
  coord_cartesian(
    xlim = c(0, 10),
    ylim = c(0, 1)
  )
```

---

# Gamma distribution

```{r gamma-5, fig.height = 6}
function_plot +
  stat_function(
    mapping = aes(color = "Gamma(3, 2)"),
    fun = dgamma, args = list(shape = 3, rate = 2), n = 501, linewidth = 1.5
  ) +
  stat_function(
    mapping = aes(color = "Gamma(5, 2)"),
    fun = dgamma, args = list(shape = 5, rate = 2), n = 501, linewidth = 1.5
  ) +
  stat_function(
    mapping = aes(color = "Gamma(3, 3)"),
    fun = dgamma, args = list(shape = 3, rate = 3), n = 501, linewidth = 1.5
  ) +
  scale_color_brewer(type = "qual") +
  xlim(0, 10) +
  labs(
    y = expression(f(x)),
    color = NULL
  ) +
  coord_cartesian(
    xlim = c(0, 10),
    ylim = c(0, 1)
  )
```

---

# Importance of the Gamma distribution

* Exponential and $\chi^2$ distributions are special cases of the gamma distribution
* Commonly used in Bayesian statistics (conjugate prior)

---

# $\chi^2$ distribution

Suppose $Z \sim \text{Normal}(0,1)$. Consider $X = Z^2$

$$\begin{aligned}
F_{X}(x)   & =  \Pr(X \leq x) \\
& = \Pr(Z^2 \leq x ) \\
& = \Pr(-\sqrt{x} \leq Z \leq \sqrt{x}) \\
& = \frac{1}{\sqrt{2\pi}} \int_{-\sqrt{x}}^{\sqrt{x} } e^{-\frac{z^2}{2}} \,dz\\
& = F_{Z} (\sqrt{x}) - F_{Z} (-\sqrt{x})
\end{aligned}$$

---

# $\chi^2$ distribution

$$
\begin{aligned}
\frac{\partial F_{X}(x) }{\partial x }  & = f_{Z} (\sqrt{x}) \frac{1}{2\sqrt{x}} + f_{Z}(-\sqrt{x}) \frac{1}{2\sqrt{x}} \\
& = \frac{1}{\sqrt{x}}\frac{1}{2 \sqrt{2\pi}} ( 2e^{-\frac{x}{2}}) \\
& = \frac{1}{\sqrt{x}}\frac{1}{\sqrt{2\pi}} ( e^{-\frac{x}{2}}) \\
& = \frac{(\frac{1}{2})^{1/2}}{\Gamma(\frac{1}{2})}\left(x^{1/2 - 1} e^{-\frac{x}{2}}\right)
\end{aligned}
$$

---

# $\chi^2$ distribution

$$f_X(x) = \frac{(\frac{1}{2})^{1/2}}{\Gamma(\frac{1}{2})}\left(x^{1/2 - 1} e^{-\frac{x}{2}}\right)$$

--

$$X \sim \text{Gamma}(1/2, 1/2)$$

--

If $X = \sum_{i=1}^{n} Z_{i}^2$, $X \sim \text{Gamma}(n/2, 1/2)$

---

# $\chi^2$ distribution

Suppose $X$ is a continuous random variable with $X\geq 0$, with PDF 

$$f(x) = \frac{1}{2^{n/2} \Gamma(n/2) } x^{n/2 - 1} e^{-x/2}$$

$X$ is a $\chi^2$ distribution with $n$ degrees of freedom

--

$$X \sim \chi^{2}(n)$$

---

# $\chi^2$ distribution: plot

```{r chi-sq, out.width="90%", fig.height = 6}
expand.grid(
  x = seq(from = 0, to = 100, length.out = 201),
  df = seq(from = 1, to = 50)
) %>%
  as_tibble() %>%
  mutate(y = dchisq(x, df)) %>%
  ggplot(mapping = aes(x = x, y = y)) +
  geom_line(linewidth = 1.5) +
  labs(
    title = "Chi-squared",
    subtitle = "{closest_state} degrees of freedom",
    y = "Density"
  ) +
  transition_states(states = df) +
  view_follow(fixed_x = TRUE)
```

---

# $\chi^2$ properties

Suppose $X \sim \chi^2(n)$

$$
\begin{aligned}
\E[X] & = \E\left[\sum_{i=1}^{N} Z_{i}^2\right] \\
 & = \sum_{i=1}^{N} \E[Z_{i}^{2} ] \\
\Var(Z_{i} ) & = \E[Z_{i}^2] - \E[Z_{i}]^2\\
1 & = \E[Z_{i}^2]- 0 \\
\E[X] & = n 
\end{aligned}
$$


---

# $\chi^2$ properties

$$
\begin{aligned}
\Var(X) & = \sum_{i=1}^{N} \Var(Z_{i}^2) \\
& = \sum_{i=1}^{N} \left(\E[Z_{i}^{4} ] - \left(\E[Z_{i}^{2}]\right)^{2}  \right) \\
& = \sum_{i=1}^{N} \left(3 - 1\right ) = 2n 
\end{aligned}
$$

---

# Student's $t$ distribution

Suppose $Z \sim \text{Normal}(0, 1)$ and $U \sim \chi^2(n)$.  Define the random variable $Y$ as, 

$$Y = \frac{Z}{\sqrt{\frac{U}{n}}}$$

--

If $Z$ and $U$ are independent then $Y \sim t(n)$, with PDF 

$$f_Y(y) = \frac{\Gamma(\frac{n+1}{2})}{\sqrt{\pi n } \Gamma(\frac{n}{2})}\left(1 + \frac{y^2}{n}\right)^{-\frac{n+1}{2}}$$

---

# Differences from the Normal distribution

* Normal distribution always has the same shape once standardized
* The shape of the student's $t$-distribution changes depending on the sample size
* At low $n$ the tails are heavier than the normal's
* As $n \uparrow$, the tails thin and the distribution tightens
* As $n \rightarrow \infty$, student's $t$-distribution takes on the same shape as the normal distribution

---

# Differences from the Normal distribution: plot

```{r t-dist, fig.height = 6, out.width="90%"}
expand.grid(
  x = seq(from = -6, 6, length.out = 201),
  df = seq(from = 1, to = 50)
) %>%
  as_tibble() %>%
  mutate(y = dt(x, df)) %>%
  ggplot(mapping = aes(x = x, y = y)) +
  geom_line(linewidth = 1.5) +
  stat_function(fun = dnorm, n = 201, color = "#0B6884") +
  labs(
    title = "Student's t",
    subtitle = "{closest_state} degrees of freedom",
    y = "Density"
  ) +
  transition_states(states = df) +
  view_follow(fixed_x = TRUE)
```

---
# HOW DO I DO THIS?!!!

--

In all cases with the continuous random variable, we are going to be using an integral OR a table (it just does the math for you!). 

You can expect to see problems where a probability distribution is given by a function (sometimes explicitly stated as a type of function (and then you need to use it) or just as the function)

---

## Example: 
After an election in a parliamentary system, a government (consisting of a prime minister and a cabinet) is formed by gathering the support of a majority of newly elected members of parliament. Typically a government is allowed to remain in power for a certain number of years before new elections must be called but elections can be called earlier. In Country Z, governments must call elections at least every 5 years, but they could be called sooner if there is a vote of no confidence or the prime minister dissolves the government. Let the continuous random variable $X$ denote the amount of time (measured in years) between the last election and the calling of the next election. $X$ has support on all real numbers between 0 and 5. Suppose we know that $X$ has the probability density function

$$f(x) = \begin{cases}
kx^3 &  0 < x < 5, \\
0 & \text{otherwise}
\end{cases}$$

where $k$ is some constant. **Find $k$.**

---

# Example: Answer

**Solution:** Recall that a valid PDF must sum up to 1. So we have to find a $k$ such that the integral of the PDF equals 1.
    
$$\begin{aligned}
\int_0^5 kx^3 dx &= 1 \\
k \int_0^5 x^3 dx &= 1 \\
k \left[ \frac{x^4}{4} \right]\bigg|_0^5 &= 1 \\
k \left[ \frac{5^4}{4} - \frac{0^4}{4} \right] &= 1 \\
\frac{625k}{4} &= 1 \\
k &= \frac{4}{625}
\end{aligned}$$
    
---

## Other question types: 

* Find the CDF
* What function is (description of scenario)
* Does the variable fall in this range? What are the chances?

---

# Recap:

  * Uniform$(a,b)$: $f(x)=\frac{1}{b-a}$; mean $\frac{a+b}{2}$, var $\frac{(b-a)^2}{12}$
  * Exponential$(\lambda)$: $f(x)=\lambda e^{-\lambda x}$; mean $\frac{1}{\lambda}$, var $\frac{1}{\lambda^2}$
  * Gamma$(\alpha,\beta)$: $f(x)=\frac{\beta^{\alpha}}{\Gamma(\alpha)}x^{\alpha-1}e^{-\beta x}$; mean $\frac{\alpha}{\beta}$, var $\frac{\alpha}{\beta^2}$
  * Normal$(\mu,\sigma^2)$: $f(x)=\frac{1}{\sqrt{2\pi\sigma^2}}e^{-(x-\mu)^2/(2\sigma^2)}$; mean $\mu$, var $\sigma^2$
  * $\chi^2(n) = $ Gamma$(n/2, 1/2)$; mean $n$, var $2n$
  * Student's $t(n)$: mean $0$ for $n>1$, var $\frac{n}{n-2}$ for $n>2$


---
class: center, middle, inverse
## Let’s get to know each other a bit more – Name, pronouns, subfield/research area, where you are currently, something fun/interesting about you and/or your hobbies