---
title: "Multivariate distributions"
author: "[MACS 33000](https://jmclip.github.io/MACSS_math_camp/) <br /> University of Chicago"
output:
  xaringan::moon_reader:
    css: ["xaringan-themer.css", "slides-menu.css"]
    math: katex
    includes:
      after_body: [slides-menu.js]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false 
---

```{r child = here::here("_common-slides.Rmd"), cache = FALSE}

```

```{r packages, include=FALSE}
library(tidyverse)
library(patchwork)
library(here)
library(gganimate)
library(plotly)
```

$$\newcommand{\E}{\mathrm{E}} \newcommand{\Var}{\mathrm{Var}} \newcommand{\Cov}{\mathrm{Cov}} \newcommand{\Cor}{\mathrm{Cor}}$$


# Misc

* Calcs spoken for
* Link for practice questions open on github / pset
* Exam rooms assigned: messaged you yesterday re: your placement
* Meetings with preceptors: Should be hearing soon re: timing


---

# Acronym recap:

* Discrete: pmf and CMF
* Continuous: pdf and CDF
  * The lowercase reflects that we're associating it with particular values (e.g. f)
  * Uppercase is integrating (e.g. F)
* Rules of probability:
  * $0 \leq p(x) \leq 1$
  * Total probability must sum to 1!
* Expected values: tells us what we can anticipate getting from our function: IN THE LONG RUN
  * multiply your value times its probability
* Moments: tell us about the function and multiply $x^n$ by the probability for the $n^{th}$ moment

---

# Learning objectives

* Define a joint probability density function
* Condition pdfs on other random variables
* Identify independence between two random variables
* Define covariance and correlation
* Examine sums of random variables
* Define the multivariate normal distribution

---

# Multivariate distribution

$X$ and $Y$ are **jointly continuous** if there exists a function $f(x,y)$ such that, for every set $C \subset \Re^{2}$,

$$\Pr\{(X, Y) \in C \}  = \iint_{(x,y)\in C} f(x,y)\, dx\, dy$$

--

What is $C \subset R^{2}$?

--

* $R^{2} = R \underbrace{\times}_{\text{Cartesian Product}} R$
* $C$ is a subset of the 2-d plane

---

## $C = \{x, y: x \in [0,1] , y\in [0,1] \}$

```{r a-2d-rect}
function_plot +
  geom_rect(data = tribble(
    ~"x1", ~"y1", ~"x2", ~"y2",
    0, 0, 1, 1
  ), mapping = aes(
    x = NULL,
    xmin = x1, xmax = x2,
    ymin = y1, ymax = y2
  ), fill = "grey") +
  xlim(-1, 1) +
  ylim(-1, 1)
```

---

# $C = \{x, y: x^2 + y^2 \leq 1 \}$

```{r a-2d-circle}
function_plot +
  ggforce::geom_circle(aes(x = NULL, x0 = 0, y0 = 0, r = 1), fill = "grey") +
  xlim(-1, 1) +
  ylim(-1, 1) +
  coord_fixed()
```

---

# $C = \{ x, y: x> y, x,y\in(0,2)\}$

```{r a-2d-triangle}
function_plot +
  geom_polygon(data = tribble(
    ~"x", ~"y",
    0, 0,
    2, 0,
    2, 2
  ), mapping = aes(x = x, y = y), fill = "grey") +
  xlim(-1, 2) +
  ylim(-1, 2)
```

---

# Multivariate distribution: role of integrals

$$\begin{aligned}
C &= \{ x, y: x \in A, y \in B \} \\
\Pr\{(X,Y) \in C \} &= \int_{B} \int_{A} f(x,y) \, dx\, dy
\end{aligned}$$

---

# Examples of joint pdfs

Consider a function $f:\Re \times \Re \rightarrow \Re$:

* Input: an $x$ value and a $y$ value.
* Output: a number from the real line
* $f(x,y) = a$


---

# Examples of joint pdfs: normal
$f(x,y) = \frac{1}{2\pi}exp[\frac{(x^2+y^2)}{-2}]$
```{r multivar-fun}
x1 <- seq(from = -2, to = 2, length.out = 101)
x2 <- seq(from = -2, to = 2, length.out = 101)

z <- expand.grid(
  x1 = x1,
  x2 = x2
) %>%
  as_tibble() %>%
  mutate(y = dnorm(x1) * dnorm(x2))
z_mat <- matrix(data = z$y, nrow = length(x1), ncol = length(x2), byrow = TRUE)

plot_ly(x = ~x1, y = ~x2, z = ~z_mat) %>%
  add_surface() %>%
  layout(
    scene = list(
      xaxis = list(title = "x"),
      yaxis = list(title = "y")
    ),
    margin = list(
      l = 5,
      r = 50,
      b = 100,
      t = 0,
      pad = 0
    )
  )
```

---

# Examples of joint pdfs: normal

```{r multivar-contour, dependson = "multivar-fun"}
ggplot(data = z, mapping = aes(x = x1, y = x2)) +
  geom_raster(mapping = aes(fill = y)) +
  scale_fill_viridis_c() +
  labs(
    x = expression(x),
    y = expression(y),
    fill = expression(f(x, y))
  )
```

---

# Examples of joint pdfs: uniform

$f(x,y) = 1$ if $x \in [0,1], y \in [0,1]$, and $f(x,y) = 0$ otherwise

```{r joint-pdf-uniform}
x1 <- seq(from = -0.5, to = 1.5, length.out = 101)
x2 <- seq(from = -0.5, to = 1.5, length.out = 101)

z <- expand_grid(
  x1 = x1,
  x2 = x2
) %>%
  mutate(y = as.numeric(x1 >= 0 & x1 <= 1 & x2 >= 0 & x2 <= 1))
z_mat <- matrix(data = z$y, nrow = length(x1), ncol = length(x2), byrow = TRUE)

plot_ly(x = ~x1, y = ~x2, z = ~z_mat) %>%
  add_surface() %>%
  layout(
    scene = list(
      xaxis = list(title = "x"),
      yaxis = list(title = "y")
    )
  )
```

---

# Examples of joint pdfs: uniform

$f(x,y) = 1$ if $x \in [0,1], y \in [0,1]$, and $f(x,y) = 0$ otherwise

```{r joint-pdf-uniform-2d, dependson = "joint-pdf-uniform"}
ggplot(data = z, mapping = aes(x = x1, y = x2)) +
  geom_raster(mapping = aes(fill = y)) +
  scale_fill_viridis_c() +
  labs(
    x = expression(x),
    y = expression(y),
    fill = expression(f(x, y))
  )
```

---

# Examples of joint pdfs: exponential

$f(x,y) = \frac{2 \exp(-2x)}{\sqrt{2\pi}}\exp\left(-\frac{(y)^2}{2}\right)$ if $x \in [0,\infty), y \in \Re$, $f(x,y) = 0$ otherwise

```{r joint-pdf-exp}
x1 <- seq(from = 0, to = 3, length.out = 101)
x2 <- seq(from = -2, to = 2, length.out = 101)

z <- expand_grid(
  x1 = x1,
  x2 = x2
) %>%
  mutate(y = dexp(x1, rate = 2) * dnorm(x2))
z_mat <- matrix(data = z$y, nrow = length(x1), ncol = length(x2), byrow = TRUE)

plot_ly(x = ~x1, y = ~x2, z = ~z_mat) %>%
  add_surface() %>%
  layout(
    scene = list(
      xaxis = list(title = "x"),
      yaxis = list(title = "y")
    )
  )
```

---

# Examples of joint pdfs

$f(x,y) = \frac{2 \exp(-2x)}{\sqrt{2\pi}}\exp\left(-\frac{(y)^2}{2}\right)$ if $x \in [0,\infty), y \in \Re$, $f(x,y) = 0$ otherwise

```{r joint-pdf-exp-2d, dependson = "joint-pdf-exp"}
ggplot(data = z, mapping = aes(x = x1, y = x2)) +
  geom_raster(mapping = aes(fill = y)) +
  scale_fill_viridis_c() +
  labs(
    x = expression(x),
    y = expression(y),
    fill = expression(f(x, y))
  )
```

---

# Examples of joint pdfs: plane

$f(x,y) = x + y$, if $x \in [0,1], y \in [0,1]$

```{r joint-pdf-plus}
x1 <- seq(from = 0, to = 3, length.out = 101)
x2 <- seq(from = 0, to = 3, length.out = 101)

z <- expand_grid(
  x1 = x1,
  x2 = x2
) %>%
  mutate(y = x1 + x2)
z_mat <- matrix(data = z$y, nrow = length(x1), ncol = length(x2), byrow = TRUE)

plot_ly(x = ~x1, y = ~x2, z = ~z_mat) %>%
  add_surface() %>%
  layout(
    scene = list(
      xaxis = list(title = "x"),
      yaxis = list(title = "y")
    )
  )
```

---

# Examples of joint pdfs: plane

$f(x,y) = x + y$, if $x \in [0,1], y \in [0,1]$

```{r joint-pdf-plus-2d, dependson = "joint-pdf-plus"}
ggplot(data = z, mapping = aes(x = x1, y = x2)) +
  geom_raster(mapping = aes(fill = y)) +
  scale_fill_viridis_c() +
  labs(
    x = expression(x),
    y = expression(y),
    fill = expression(f(x, y))
  )
```

---
class: middle

# Evaluating probability

--
### IT WORKS THE SAME-*ish*

---

# Multivariate CDF

For jointly continuous random variables $X$ and $Y$ define, $F(b,a)$ as

$$F(b,a) = \Pr\{ X \leq b , Y \leq a\} = \int_{-\infty}^{a} \int_{-\infty}^{b} f(x,y) \, dx\, dy$$

---

# Multivariate CDF

```{r multi-cdf}
x1 <- seq(from = -1.5, to = 1.5, length.out = 101)
x2 <- seq(from = -1.5, to = 1.5, length.out = 101)

z <- expand.grid(
  x1 = x1,
  x2 = x2
) %>%
  as_tibble() %>%
  mutate(y = 6*x1^2 + x2^2)

multi_cdf <- ggplot(data = z, mapping = aes(x = x1, y = x2)) +
  geom_raster(mapping = aes(fill = y)) +
  scale_fill_viridis_c() +
  labs(
    x = expression(x),
    y = expression(y),
    fill = expression(f(x, y))
  )
multi_cdf
```

---

# Multivariate CDF

```{r multi-cdf-1, dependson = "multi-cdf"}
multi_cdf +
  geom_rect(
    data = tibble(x = -1, y = -0.5),
    mapping = aes(
      xmin = x, ymin = y, xmax = -1.5, ymax = -1.5,
      x = NULL, y = NULL
    ),
    color = "black", alpha = 0.5
  ) +
  coord_cartesian(xlim = c(-1.5, 1.5)) +
  ggtitle(expression(F(-1, -0.5)))
```

---

# Multivariate CDF

```{r multi-cdf-2, dependson = "multi-cdf"}
multi_cdf +
  geom_rect(
    data = tibble(x = -0.5, y = -1.25),
    mapping = aes(
      xmin = x, ymin = y, xmax = -1.5, ymax = -1.5,
      x = NULL, y = NULL
    ),
    color = "black", alpha = 0.5
  ) +
  coord_cartesian(xlim = c(-1.5, 1.5)) +
  ggtitle(expression(F(-0.5, -1.25)))
```

---

# Multivariate CDF

```{r multi-cdf-3, dependson = "multi-cdf"}
multi_cdf +
  geom_rect(
    data = tibble(x = 0, y = 0),
    mapping = aes(
      xmin = x, ymin = y, xmax = -1.5, ymax = -1.5,
      x = NULL, y = NULL
    ),
    color = "black", alpha = 0.5
  ) +
  coord_cartesian(xlim = c(-1.5, 1.5)) +
  ggtitle(expression(F(0, 0)))
```

---

# Multivariate CDF

```{r multi-cdf-4, dependson = "multi-cdf"}
multi_cdf +
  geom_rect(
    data = tibble(x = 1.5, y = 0.25),
    mapping = aes(
      xmin = x, ymin = y, xmax = -1.5, ymax = -1.5,
      x = NULL, y = NULL
    ),
    color = "black", alpha = 0.5
  ) +
  coord_cartesian(xlim = c(-1.5, 1.5)) +
  ggtitle(expression(F(1.5, 0.25)))
```

---
class: middle

# MARGINAL PROBABILITIES!

--
Remember how we talked about $f_X$? ....
--
Now we can see why/how that would be worthwhile. (how one variable changes relative to another vs on its own)

---

# Marginalization

Define $f_{X}(x)$ as the pdf for $X$, 

$$f_{X}(x) = \int_{-\infty}^{\infty} f(x,y) \, dy$$

--

Similarly, define $f_{Y}(y)$ as the pdf for $Y$,

$$f_{Y}(y) = \int_{-\infty}^{\infty} f(x,y)\, dx$$

--

$X$ and $Y$ are continuous random variables with joint pdf $f(x,y)$. Define the conditional probability function $f(x|y)$ as 

$$f(x|y) = \frac{f(x, y) }{f_{Y}(y) }$$

---

# Marginalization in action
$f(x,y) = -x^2 - y^2$, if $x \in [-2,2], y \in [-2,2]$

```{r joint-pdf-EXmar}
x1 <- seq(from = -2, to = 2, length.out = 101)
x2 <- seq(from = -2, to = 2, length.out = 101)

z <- expand_grid(
  x1 = x1,
  x2 = x2
) %>%
  mutate(y = -x1^2 - x2^2)
z_mat <- matrix(data = z$y, nrow = length(x1), ncol = length(x2), byrow = TRUE)

plot_ly(x = ~x1, y = ~x2, z = ~z_mat) %>%
  add_surface() %>%
  layout(
    scene = list(
      xaxis = list(title = "x"),
      yaxis = list(title = "y")
    )
  )
```

---

# Marginalization

$$f(x, y) = -x^2 - y^2$$

.pull-left[

```{r conditional-pdf, fig.width = 6}
x1 <- seq(from = -2, to = 2, length.out = 201)
x2 <- seq(from = -2, to = 2, length.out = 201)

z <- expand_grid(
  x1 = x1,
  x2 = x2
) %>%
  mutate(y = dnorm(x1) * dnorm(x2))

ggplot(data = z, mapping = aes(x = x1, y = x2)) +
  geom_raster(mapping = aes(fill = y)) +
  scale_fill_viridis_c() +
  labs(
    x = expression(x),
    y = expression(y),
    fill = expression(f(x, y))
  ) +
  coord_fixed()
```

]

--

.pull-right[

```{r conditional-pdf-examples, fig.width = 6}
x1 <- seq(from = -2, to = 2, length.out = 50)
x2 <- seq(from = -2, to = 2, length.out = 50)

z <- expand_grid(
  x1 = x1,
  x2 = x2
) %>%
  mutate(y = dnorm(x1) * dnorm(x2))

function_plot +
  geom_line(
    data = z,
    mapping = aes(x = x1, y = y, color = y), linewidth = 6
  ) +
  scale_color_viridis_c() +
  labs(
    title = expression(f(x, y) == -x^2 - y^2),
    subtitle = "y = {closest_state}",
    x = expression(x),
    y = expression(z)
  ) +
  transition_states(states = x2, wrap = FALSE) +
  view_follow()
```

]

---

# Why does marginalization work?

### Discrete case

Consider jointly distributed discrete random variables, $X$ and $Y$ with joint PMF

$$\Pr(X =x, Y = y) = p(x, y)$$

--

Suppose that the distribution allocates its mass at $x_{1}, x_{2}, \ldots, x_{M}$ and $y_{1}, y_{2}, \ldots, y_{N}$. Define the conditional mass function $\Pr(X= x| Y= y)$ as,

$$\begin{aligned}
\Pr(X=x|Y=y) &\equiv p(x|y)  \\
& = \frac{p(x,y)}{p(y)}
\end{aligned}$$

--

$$p(x,y) = p(x|y)p(y)$$

---
# Marginalization (discrete):
So, we have: $$p(x,y) = p(x|y)p(y)$$

--

Marginalizing **over** $y$ to get $p(x)$ is then,

$$p(x_{j}) = \sum_{i=1}^{N} p(x_{j} |y_{i})p(y_{i} )$$

--

Remember when we were thinking about our probability across cases? Same logic!

---

# Why does marginalization work?

|  | $Y = 0$ | $Y = 1$ |  |
|-------|-------------|-------------|------------|
| $X = 0$ | $p(0,0)$ | $p(0, 1)$ | $p_{X}(0)$ |
| $X = 1$ | $p(1,0)$ | $p(1,1)$ | $p_{X}(1)$ |
|  | $p_{Y} (0)$ | $p_{Y} (1)$ |  |


---

# Why does marginalization work?

|  | $Y = 0$ | $Y = 1$ |  |
|-------|-------------|-------------|------------|
| $X = 0$ | $0.01$ | $0.05$ | $p_{X}(0)$? |
| $X = 1$ | $0.25$ | $0.69$ | $p_{X}(1)$? |
|  | $0.26$ | $0.74$ |  |

--

$$p_{X}(0) = \Pr(0|y = 0) *\Pr(y= 0) + \Pr(0|y=1)*\Pr(y=1)$$

--

$$= \frac{0.01}{0.26} * 0.26 + \frac{0.05}{0.74} * 0.74 \\
 = 0.06$$

--

$$\begin{aligned}
p_{X}(1) & = \Pr(1|y = 0) \Pr(y= 0) + \Pr(1|y=1) \Pr(y=1) \\ 
& = \frac{0.25}{0.26} * 0.26 + \frac{0.69}{0.74} * 0.74 \\
& = 0.94
\end{aligned}$$

---

# Why does marginalization work?

|  | $Y = 0$ | $Y = 1$ |  |
|-------|-------------|-------------|------------|
| $X = 0$ | $0.01$ | $0.05$ | $0.06$|
| $X = 1$ | $0.25$ | $0.69$ | $0.94$ |
|  | $0.26$ | $0.74$ |  |

---

# Why does marginalization work?

### Continuous case

For **jointly distributed continuous** random variables $X$ and $Y$ define,

$$f_{X|Y}(x|y) = \frac{f(x,y)}{f_{Y}(y) }$$

--

$$f_{X}(x) = \int_{-\infty }^{\infty} f_{X|Y}(x|y)f_{Y}(y) \, dy$$

--

* Think of $f_{X|Y}(x|y)$ as the pdf for $X$ at a value of $Y$
* Average over those pdfs to get the final pdf for $X$
    * Weighted average

---

# A simple(ish) example

Suppose $X$ and $Y$ are jointly continuous and that

$$
\begin{aligned}
f(x,y) &  = x + y \text{ , if  } x \in [0,1], y \in [0,1] \\
& = 0 \text{ , otherwise } 
\end{aligned}
$$

```{r joint-pdf-plus-dup, fig.height = 6}
x1 <- seq(from = 0, to = 1, length.out = 101)
x2 <- seq(from = 0, to = 1, length.out = 101)

z <- expand_grid(
  x1 = x1,
  x2 = x2
) %>%
  mutate(y = x1 + x2)

ggplot(data = z, mapping = aes(x = x1, y = x2)) +
  geom_raster(mapping = aes(fill = y)) +
  scale_fill_viridis_c() +
  labs(
    x = expression(x),
    y = expression(y),
    fill = expression(f(x, y))
  )
```

---

# A simple(-ish) example

We want $f_{X}(x)$. We can find this one of two ways: by either taking the derivative of the original function with respect to y OR by using independence. 

Option 1: $\int_0^1f(x,y) dy = \int_0^1 (x+ y) dy = xy + \frac{y^2}{2} |_0^1 = x + 1/2$

--

Option 2: we know that $f(x|y) = \frac{f(x,y)}{f_Y(y)}$

--

Suppose you are given: $f_{Y}(y) = 1/2 + y$

--

Then

$$f(x|y) = \frac{x + y}{1/2 + y}$$

$$f(x) = \int_{0}^{1} f(x|y)f(y)\, dy = 1/2 + x$$

---

# A simple(-ish) example

Here is $f(x \mid y)$ -- one curve for each value of $y$:

```{r joint-pdf-plus-margin}
conditional_pdf <- function(x, y) {
  (x + y) / (1 / 2 + y)
}

expand_grid(
  x = seq(from = 0, to = 1, length.out = 101),
  y = seq(from = 0, to = 1, by = .001)
) %>%
  mutate(cond_x = conditional_pdf(x, y)) %>%
  ggplot(mapping = aes(x, cond_x, color = y, group = y)) +
  geom_line() +
  scale_color_viridis_c("Conditioning \n value y", option = "plasma") +
  labs(
    x = expression(x),
    y = expression(f(x ~ "|" ~ y)),
    color = expression(y)
  )
```

---

# A simple(-ish) example
Here's our 'weighted average' for f(x) across all y: 
```{r joint-pdf-plus-cond, dependson = "joint-pdf-plus-margin"}
expand_grid(
  x = seq(from = 0, to = 1, length.out = 101),
  y = seq(from = 0, to = 1, by = .001)
) %>%
  mutate(cond_x = conditional_pdf(x, y)) %>%
  ggplot(mapping = aes(x, cond_x, color = y, group = y)) +
  geom_line(alpha = .10) +
  geom_abline(intercept = 0.5, slope = 1) +
  scale_color_viridis_c("Conditioning \n value y", option = "plasma") +
  labs(
    x = expression(x),
    y = expression(f(x)),
    color = expression(y)
  )
```

---

# More complex example

Suppose $X$ and $Y$ are jointly distributed with pdf 

$$f(x,y) = 2 \exp(-x) \exp(-2y), \forall \, x>0, y>0$$

--

Note: equivalent to :
$$f(x,y) = 2 e^{(-x)} e^{(-2y)}, \forall \, x>0, y>0$$

---

# Verify this is a pdf (e version)

$$\int_{0}^{\infty} \int_{0}^{\infty} f(x, y) \, dx\, dy = 2\int_{0}^{\infty} \int_{0}^{\infty} e^{(-x)} e^{(-2y)} \, dx\, dy$$

--

$$2 \int_{0}^{\infty}e^{(-2y)} \, dy \int_{0}^{\infty} e^{(-x)} \, dx$$
--
$$= 2 (-\frac{1}{2} e^{(-2y)}|_{0}^{\infty}  ) ( - e^{(-x)}|_{0}^{\infty} )$$
--
$$= 2\left[ (-\frac{1}{2}(\lim_{y\rightarrow\infty} e^{(-2y)} - 1))(- (\lim_{x\rightarrow \infty} e^{(-x)} - 1) ) \right]$$
--
$$= 2 \left[  -\frac{1}{2} (-1) * -1 (-1)   \right] = 1$$
---

# Verify this is a pdf (exp version)

$$\int_{0}^{\infty} \int_{0}^{\infty} f(x, y) \, dx\, dy = 2\int_{0}^{\infty} \int_{0}^{\infty} \exp(-x) \exp(-2y) \, dx\, dy$$
--

$$2\int_{0}^{\infty}\exp(-2y) \, dy \int_{0}^{\infty} \exp(-x) \, dx$$
--
$$= 2 (-\frac{1}{2} \exp(-2y)|_{0}^{\infty}  ) ( - \exp(-x)|_{0}^{\infty} )$$
--
$$= 2\left[ (-\frac{1}{2}(\lim_{y\rightarrow\infty} \exp(-2y) - 1))(- (\lim_{x\rightarrow \infty} \exp(-x) - 1) ) \right]$$
--
$$= 2 \left[  -\frac{1}{2} (-1) * -1 (-1)   \right]= 1$$

---

# Calculate CDF

$$F(b,a) \equiv \Pr\{X \leq b, Y \leq a\} = 2 \int_{0}^{a} \int_{0}^{b} \exp(-x) \exp(-2y) \, dx\, dy$$
--
$$= 2 (\int_{0}^{a} \exp(-2y) \, dy) (\int_{0}^{b} \exp(-x) \, dx)$$
--
$$ = 2 \left[-\frac{1}{2} (\exp(-2a) -1 )\right]\left[ - (\exp(-b) - 1) \right]$$

--
$$ = \left[1 - \exp(-2a) \right] \left[ 1- \exp(-b) \right]$$
--
NOW WHAT? 
--
Plug in values for a and b, respectively. 

---

# Calculate $f_{X}(x)$ and $f_{Y}(y)$ 

$$\begin{aligned}f_{X}(x) & = \int_{0}^{\infty} 2\exp(-x) \exp(-2y) \, dy \\& = 2 \exp(-x) \int_{0}^{\infty} \exp(-2y) \, dy \\& = 2 \exp(-x) \left[ -\frac{1}{2}(0 - 1) \right] \\& = \exp(-x)\end{aligned}$$

--

$$\begin{aligned}f_{Y}(y) & = \int_{0}^{\infty} 2 \exp(-x) \exp(-2y) \, dx \\& = 2 \exp(-2y) \int_{0}^{\infty} \exp(-x) \, dx \\& = 2 \exp(-2y) \left[-(0 -1) \right] \\& = 2 \exp(-2y)\end{aligned}$$

---

# Independence

Two random variables $X$ and $Y$ are independent if for any two sets of real numbers $A$ and $B$,  

$$\Pr\{ X \in A , Y \in B \} = \Pr\{X \in A\} \Pr\{Y \in B\}$$

--

$$f(x,y) = f_{X}(x) f_{Y}(y)$$

--

If $X$ and $Y$ are not independent, we will say they are **dependent**.

--

If $X$ and $Y$ are independent, then

$$\begin{aligned}f_{X|Y} (x|y) & = \frac{f(x,y)}{f_{Y}(y)} \\& = \frac{f_{X}(x)f_{Y}(y)}{f_{Y}(y) } \\& = f_{X}(x) \end{aligned}$$

---

# A (simple) example of dependence

Suppose $X$ and $Y$ are jointly continuous and that

$$\begin{aligned}f(x,y) &  = & x + y \text{ , if  } x \in [0,1], y \in [0,1] \\& = & 0 \text{ , otherwise }  \end{eqnarray}$$

--

Are they dependent? 

---

# A (simple) example of dependence


$$f_{X}(x) = \int_{0}^{1} \left(x + y \right) \, dy$$
--
$$xy  + \frac{y^2}{2} |^{1}_{0} = x + \frac{1}{2}$$
--
$$f_{Y}(y) = \int_{0}^{1} \left(x + y \right) \, dx =  \frac{x^2}{2} + xy  |^{1}_{0} = \frac{1}{2} + y$$
--
$$\begin{aligned}f_{X}(x) f_{Y}(y) & = & (\frac{1}{2} + x) (\frac{1}{2} + y) \\& = & \frac{1}{4} + \frac{x + y}{2} + xy \neq (x+y)\end{eqnarray}$$
---
# Dependence: deep thoughts

$$\begin{aligned}f_{X}(x) f_{Y}(y) & = & (\frac{1}{2} + x) (\frac{1}{2} + y) \\& = & \frac{1}{4} + \frac{x + y}{2} + xy \neq (x+y)\end{eqnarray}$$

We can think about this as getting additional INFORMATION from y when we know x and vice versa. 

---
class: middle

# But what to expect?

---

# Expectation

For jointly continuous random variables $X$ and $Y$,

--

$$\E[X] = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} x f(x,y) \, dx\, dy$$
--

$$E[Y] = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} y f(x,y) \, dx\, dy$$
--
$$E[XY] = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} x y f(x,y) \, dx\, dy$$

--

Suppose $g:\Re^{2} \rightarrow \Re$. Then

$$\E[g(X, Y)] = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} g(x, y) f(x,y) \, dx\, dy$$
---

# Expectation ex with g(x)

$$\E[g(X, Y)] = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} g(x, y) f(x,y) \, dx\, dy$$

Suppose we have the following functions: $f(x,y) = xy$ where $0 \leq x \leq 1$ and $0 \leq y \leq 2$. 
1. Is this a valid pdf? why or why not?
2. Suppose you earn revenue equal to $x^3+2$. What is your expected utility? 

---

# Expectation ex with g(x): pdf


Suppose we have the following functions: $f(x,y) = xy$ where $0 \leq x \leq 1$ and $0 \leq y \leq 2$. 

$$\int_0^2\int_0^1 xy\, \,dx\,dy$$
--
$$\int_0^2\frac{x^2y}{2}|_0^1 \, dy= \int_0^2 \frac{y}{2} \, dy$$
$$=\frac{y^2}{4}|_0^2 = 1$$

--
(also want to check the value over our range of evaluation)
So, yes it appears to satisfy our requirements!


---

# Expectation ex with g(x): expectation
 > Suppose you earn revenue equal to $x^3+2$. What is your expected revenue? 

Well, our function is: $\int_0^2\int_0^1 xy\, \,dx\,dy$, so our expected revenue will be: 
$$\int_0^2\int_0^1 (x^3+2)*xy\, \,dx\,dy$$

We know we can do this because: 
$$\E[g(X, Y)] = \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} g(x, y) f(x,y) \, dx\, dy$$
---
# Expected value: Math party

$$\int_0^2\int_0^1 (x^3+2)*xy\, \,dx\,dy = \int_0^2\int_0^1 (x^4+2x)y\, \,dx\,dy$$
--
$$=\int_0^2 y \,dy* \int_0^1 (x^4+2x) \,dx= \frac{y^2}{2}|_0^2*(\frac{x^5}{5}+x^2 )|_0^1$$
--
$$=2*(1^5/5+1)=2.4$$
--
NOW WHAT? 

--
This is our expected value!


---
class: center, middle, inverse

# Examples!

---
# Example land: 

Let's look at what we have here: $\int_{1}^{2} \int_{0}^{0.5} (x^2 - \frac{8y}{3}+\frac{1}{3}) \text{ } dy \text{ }dx$

```{r multi-cdf-int, }
x1 <- seq(from = -1.5, to = 2, length.out = 101)
x2 <- seq(from = -1.5, to = 1.5, length.out = 101)

z <- expand.grid(
  x1 = x1,
  x2 = x2
) %>%
  as_tibble() %>%
  mutate(y = x1^2 - (8/3)*x2+1/3)

multi_cdf2 <- ggplot(data = z, mapping = aes(x = x1, y = x2)) +
  geom_raster(mapping = aes(fill = y)) +
  scale_fill_viridis_c() +
  labs(
    x = expression(x),
    y = expression(y),
    fill = expression(f(x, y))
  )

multi_cdf2 +
  geom_rect(
    data = tibble(x = 2, y = 0.5),
    mapping = aes(
      xmin = x, ymin = y, xmax = 1, ymax = 0,
      x = NULL, y = NULL
    ),
    color = "black", alpha = 0.5
  ) +
  coord_cartesian(xlim = c(-1.5, 2)) +
  ggtitle(expression(F(2, 0.5)))
```

---
# Example land: 

Let's look at what we have here:  $\int_{1}^{2} \int_{0}^{0.5} (x^2 - \frac{8y}{3}+\frac{1}{3}) \text{ } dy \text{ }dx$

So we will effectively add up all the chunks of our function in one direction, then we can add up along the other. I have this set up for doing y first. I leave doing it in the x direction as an exercise. 

--

$$\begin{align}F(x,y) &= \int_{1}^{2} \int_{0}^{0.5} (x^2 - \frac{8y}{3}+\frac{1}{3}) \text{ } dy \text{ }dx \\ &=\int_{1}^{2}(x^2y - \frac{4y^2}{3}+\frac{y}{3}) \text{ } |_0^{0.5} \text{ }dx \\
&= \int_{1}^{2}(0.5x^2 - \frac{1}{3}+\frac{1}{6}) \text{ } \text{ }dx \\
&= \int_{1}^{2}(0.5x^2 - \frac{1}{6}) \text{ } \text{ }dx \\
&= (\frac{x^3}{6} - \frac{x}{6}) \text{ }|_1^2 = (\frac{8}{6}- \frac{2}{6})-(\frac{1}{6}-\frac{1}{6}) \text{ } \\
&= 1
\end{align}$$
--
PHEW! It's a pdf (also need to verify we don't have negative values for f(x)) (it is: the minimum over $x \in [1,2], y \in [0,0.5]$ is at $(1, 0.5)$, where $f = 1 - \frac{4}{3} + \frac{1}{3} = 0$.)


---
# Example land: 

Let's look at a slice of it:  $\int_{1}^{1.5} \int_{0}^{0.5} (x^2 - \frac{8y}{3}+\frac{1}{3}) \text{ } dy \text{ }dx$

$$\begin{align}F(x,y) &= \int_{1}^{1.5} \int_{0}^{0.5} (x^2 - \frac{8y}{3}+\frac{1}{3}) \text{ } dy \text{ }dx \\ &=\int_{1}^{1.5}(x^2y - \frac{4y^2}{3}+\frac{y}{3}) \text{ } |_0^{0.5} \text{ }dx \\
&= \int_{1}^{1.5}(0.5x^2 - \frac{1}{3}+\frac{1}{6}) \text{ } \text{ }dx \\
&= \int_{1}^{1.5}(0.5x^2 - \frac{1}{6}) \text{ } \text{ }dx \\
&= (\frac{x^3}{6} - \frac{x}{6}) \text{ }|_1^{1.5} \\
&= (\frac{1.5^3}{6}-\frac{3}{12})-(\frac{1}{6}- \frac{1}{6}) \text{ } \\
&= 0.3125
\end{align}$$
--
So....what does this tell us??

---
class: middle

# And now, something quite a bit different*
*(but also not!)*

---

# Covariance and correlation
We are interested in how our variables vary with one another. 
For jointly continuous random variables $X$ and $Y$ we define the covariance of $X$ and $Y$ as,

$$Cov(X,Y) = E[(X-E[X])(Y- E[Y])]$$
--
$$=E\left[ XY - E[X] Y - E[Y] X + E[X]E[Y] \right]$$
--
$$=E[XY] - 2E[X]E[Y] + E[E[X]E[Y]]$$
--
$$=E[XY] - E[X]E[Y]$$

--

Define the correlation of $X$ and $Y$ as,

$$\Cor(X,Y) =  \frac{\Cov(X,Y) }{\sqrt{\Var(X) \Var(Y) } }$$

---

# Variance is covariance with itself

$$\begin{aligned}\Cov(X,X) & = & E[X X] - E[X]E[X] \\& = & E[X^2] - E[X]^2\end{eqnarray}$$

---

# Correlation measures linear relationship
Like covariance, bigger is more, but now we're standardizing relative to variance. 
Suppose $X = Y$:

$$\begin{aligned}\Cor(X,Y) & = & \frac{\Cov(X,Y)}{\sqrt{\Var(X)\Var(Y)} } \\& = & \frac{\Var(X)}{\Var(X)} \\& = & 1\end{eqnarray}$$

--

Suppose $X = -Y$:

$$\begin{aligned}\Cor(X,Y) & = & \frac{\Cov(X,Y)}{\sqrt{\Var(X)\Var(Y)} } \\& =  & \frac{- \Var(X)}{\Var(X)} \\& = & -1 \end{eqnarray}$$

---

# Correlation is between -1 and 1

$$|\Cor(X,Y)| \leq 1$$

---

# Sums of random variables

Suppose we have a sequence of random variables $X_{i}$ , $i = 1, 2, \ldots, N$

Suppose that they have joint pdf

$$f(\boldsymbol{x}) = f(x_{1}, x_{2}, \ldots, x_{n})$$

--

* $\E[\sum_{i=1}^{N}X_{i} ]  = \sum_{i=1}^{N} \E[X_{i}]$

--

* $\Var(\sum_{i=1}^{N} X_{i} )   = \sum_{i=1}^{N} \Var(X_{i} )  + 2 \sum_{i<j} \Cov(X_{i}, X_{j})$ 

---

# Expected value of sums of RVs

Suppose we have a sequence of random variables $X_{i}$ , $i = 1, 2, \ldots, N$

Suppose that they have joint pdf,

$$f(\boldsymbol{x}) = f(x_{1}, x_{2}, \ldots, x_{n})$$

Then 

$$\E[\sum_{i=1}^{N} X_{i} ] = \sum_{i=1}^{N} \E[X_{i} ]$$

--

$$
\begin{aligned}
\E[\sum_{i=1}^{N} X_{i} ] & = &  \E[X_{1} + X_{2} + \ldots + X_{N}] \\
& = & \int_{-\infty}^{\infty} \cdot \cdot \cdot \iint_{-\infty}^{\infty} (x_{1} + x_{2} + \ldots + x_{N}) f(x_{1}, x_{2}, \ldots, x_{N}) \, dx_{1}\, dx_{2}\ldots \, dx_{N} \\
& = & \int_{-\infty}^{\infty}x_{1} f_{X_{1}}(x_{1}) \, dx_{1}  + \int_{-\infty}^{\infty}x_{2} f_{X_{2}}(x_{2}) \, dx_{2} + \ldots + \int_{-\infty}^{\infty}x_{N} f_{X_{N}}(x_{N}) \, dx_{N}  \\
& = & \E[X_{1} ] + \E[X_{2}] + \ldots + \E[X_{N}] 
\end{eqnarray}
$$

---

# Variance of sums of RVs

Suppose $X_{i}$ is a sequence of random variables.  Then 

$$
\begin{aligned}
\Var(\sum_{i=1}^{N} X_{i} ) & = & \sum_{i=1}^{N} \Var(X_{i} )  + 2 \sum_{i<j} \Cov(X_{i}, X_{j} ) 
\end{eqnarray}
$$

--

Consider two random variables, $X_{1}$ and $X_{2}$.  Then, 

$$
\begin{aligned}
\Var(X_{1} + X_{2} ) & = & \E[(X_{1} + X_{2})^2] - \left(\E[X_{1}] + \E[X_{2}] \right)^2 \\ 
& = & \E[X_{1}^2]  + 2 \E[X_{1}X_{2}]  + \E[X_{2}^2]  \\
&& - (\E[X_{1}])^2 - 2 \E[X_{1}] \E[X_{2}]  - 2 \E[X_{2}]^2 \\
& = & \underbrace{\E[X_{1}^2] - (\E[X_{1}])^2}_{\Var(X_{1}) }   + \underbrace{\E[X_{2}^2] - \E[X_{2}]^{2}}_{\Var(X_{2})} \\
&&  + 2 \underbrace{(\E[X_{1} X_{2} ]  - \E[X_{1}] \E[X_{2} ] )}_{\Cov(X_{1}, X_{2} ) } \\
& = & \Var(X_{1} ) + \Var(X_{2} ) + 2 \Cov(X_{1}, X_{2})
\end{eqnarray}
$$

---
class: middle

# Oh good, matrices!
--

### AKA: What happens when you bring multiple variables to the party? 
Note: This portion is informational only! 

--

These are for information -- they haven't featured into my work, but for some people they have been relevant 

---

# Multivariate normal distribution

Suppose $\boldsymbol{X} = (X_{1}, X_{2}, \ldots, X_{N})$ is a vector of random variables.  If $\boldsymbol{X}$ has pdf 

$$f(\boldsymbol{x}) = (2 \pi)^{-N/2} \text{det}\left(\boldsymbol{\Sigma}\right)^{-1/2} \exp\left(-\frac{1}{2}(\boldsymbol{x} - \boldsymbol{\mu})^{'}\boldsymbol{\Sigma}^{-1} (\boldsymbol{x} - \boldsymbol{\mu} ) \right)$$
where $det(\mathbf{\Sigma})$ is the determinant of the covariance matrix, known as generalized variance

--

$\boldsymbol{X}$ is a **Multivariate Normal** Distribution,

$$\boldsymbol{X} \sim \text{Multivariate Normal} (\boldsymbol{\mu}, \boldsymbol{\Sigma})$$

--

* Regularly used for likelihood, Bayesian, and other parametric inferences

---

# Bivariate example

Consider the (bivariate) **special case** where $\boldsymbol{\mu} = (0, 0)$ and 

$$\boldsymbol{\Sigma} = \begin{pmatrix} 1 & 0 \\0 & 1 \\\end{pmatrix}$$

--

$$\begin{aligned}f(x_{1}, x_{2} ) & = & (2\pi)^{-2/2} 1^{-1/2} \exp\left(-\frac{1}{2}\left( (\boldsymbol{x}  - \boldsymbol{0} ) ^{'} \begin{pmatrix} 1 & 0 \\0 & 1 \\\end{pmatrix}(\boldsymbol{x}  - \boldsymbol{0} ) \right) \right) \\& = & \frac{1}{2\pi} \exp\left(-\frac{1}{2} (x_{1}^{2} + x_{2} ^ 2 )     \right) \\& = & \frac{1}{\sqrt{2\pi}} \exp\left(-\frac{x_{1}^{2}}{2}  \right) \frac{1}{\sqrt{2\pi}} \exp\left(-\frac{x_{2}^{2}}{2}  \right)\end{eqnarray}$$

--

$\leadsto$ product of univariate standard normally distributed random variables

---

# Standard multivariate normal

Suppose $\boldsymbol{Z} = (Z_{1}, Z_{2}, \ldots, Z_{N})$ is 

$$\boldsymbol{Z} \sim \text{Multivariate Normal}(\boldsymbol{0}, \boldsymbol{I}_{N} )$$

Then we will call $\boldsymbol{Z}$ the standard multivariate normal

---

# Properties of the multivariate normal

Suppose $\boldsymbol{X} = (X_{1}, X_{2}, \ldots, X_{N} )$

$$
\begin{aligned}
\E[\boldsymbol{X} ] & = &  \boldsymbol{\mu} \\
\Cov(\boldsymbol{X} ) & = & \boldsymbol{\Sigma} 
\end{eqnarray}
$$

So that, 

$$
\begin{aligned}
\boldsymbol{\Sigma}  & = & \begin{pmatrix} 
\Var(X_{1}) & \Cov(X_{1}, X_{2}) & \ldots & \Cov(X_{1}, X_{N}) \\
\Cov(X_{2}, X_{1}) & \Var(X_{2}) & \ldots & \Cov(X_{2}, X_{N} ) \\
\vdots & \vdots & \ddots & \vdots \\
\Cov(X_{N}, X_{1} ) & \Cov(X_{N}, X_{2} ) & \ldots & \Var(X_{N} ) \\
\end{pmatrix} 
\end{eqnarray}
$$

---

# Independence and multivariate normal

Suppose $X$ and $Y$ are independent.  Then

$$\Cov(X, Y) = 0$$

--


$$\Cov(X, Y) = \E[XY] - \E[X]\E[Y]$$

--

$$
\begin{aligned}
\E[XY] & = & \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} x y f(x,y)\, dx\, dy \\
& =& \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} x y f_{X}(x) f_{Y}(y)\, dx\, dy \\
& = & \int_{-\infty}^{\infty} x f_{X}(x) \, dx \int_{-\infty}^{\infty} y f_{Y}(y) \, dy \\
& = & \E[X] \E[Y]
\end{eqnarray}
$$

--

Then $\Cov(X,Y) = 0$.

---

# Independence and multivariate normal

**Zero covariance does not generally imply independence!**

--

* Suppose $X \in \{-1, 1\}$ with $\Pr(X = 1) = \Pr(X = -1) = 1/2$
* Suppose $Y \in \{-1, 0,1\}$ with $Y = 0$ if $X = -1$ and $\Pr(Y = 1) = \Pr(Y= -1)$ if $X = 1$

--

$$
\begin{aligned}
\E[XY] & = & \sum_{i \in \{-1, 1\} } \sum_{j \in \{-1, 0, 1\}} i j \Pr(X = i, Y = j) \\
& = & -1 * 0 *  \Pr(X = -1, Y = 0) + 1 * 1  * \Pr(X = 1, Y = 1)  \\
 && - 1 * 1 * \Pr(X = 1, Y = -1) \\
 &= & 0 + \Pr(X = 1, Y = 1)  - \Pr(X = 1, Y = -1 ) \\
  & = &  0.25 - 0.25 = 0  \\
\E[X] & = &  0 \\
\E[Y] & = & 0 
\end{eqnarray}
$$
--

But $Y$ is completely determined by $X$: knowing $X = -1$ tells you $Y = 0$ with certainty. So $\Cov(X,Y) = 0$ while $X$ and $Y$ are **dependent**.


---
# Independence, matrices, covariance: why might we care?

* we are looking for information
* duplication/dependence in variables can be problematic for predicting relationships (recall: determinants and eigenvalues)


---
# Recap!

* Joint pdf: same concept, integrate over variables
* Condition pdfs on other random variables and marginal probability
* Covariance and correlation: linear relationship
* Matrices: can keep going with many dimensions!

---
class: center, middle, inverse
## Let’s get to know each other a bit more – Name, pronouns, subfield/research area, where you are currently, something fun/interesting about you and/or your hobbies
